---
title: "Supplementary Figures"
output: html_notebook
---

```{r}
rm(list = setdiff(ls(), "so_all"))
dir_proj <- "/singerlab/linglin/Th17_single_cell_eae_ut/"

library(ggplot2)
library(RColorBrewer)
library(gridExtra)
library(grid)
library(cowplot)
library(dplyr)
library(tidyr)
library(tibble)
library(ggpubr)
library(Seurat)
library(aricode)
library(pheatmap)

source(paste0(dir_proj, "1_code/utils.R"))
```

Load data.
```{r, eval=FALSE}
prep_type_date <- "dominant_TCR_2020-03-24"
so_all <- readRDS(paste0(dir_proj, "2_pipeline/preprocessing/so_processed_", prep_type_date, ".rds"))
```



# Figure S1
```{r}
dir_out <- paste0(dir_proj, "3_output/Figure_S1/")
if(!dir.exists(dir_out)) dir.create(dir_out, recursive = T)
```

## Figure S1D, G: UMAP showing clusters (D) and tissue batches (G)
```{r}
date_cluster <- "2020-04-13"
reductions <- readRDS(paste0(dir_proj, "2_pipeline/clustering/UT_GFPpos_inter/", date_cluster, "/FILES/reductions.rds"))
meta <- read.table(paste0(dir_proj, "2_pipeline/clustering/UT_GFPpos_inter/", date_cluster, "/FILES/meta_data.csv"), sep = ",", header = T, row.names = 1)
stopifnot(all(rownames(reductions$umap@cell.embeddings) == rownames(meta)))
meta$UMAP_1 <- reductions$umap@cell.embeddings[,"UMAP_1"]
meta$UMAP_2 <- reductions$umap@cell.embeddings[,"UMAP_2"]

## S1D: cluster
p <- ggplot(meta, aes(x = UMAP_1, y = UMAP_2, color = factor(seurat_clusters))) + 
  geom_point(size = 0.1) +
  labs(color = "Cluster") +
  guides(color = guide_legend(override.aes = list(size = 5))) +
  theme_classic() +
  theme(legend.title = element_text(size = 15),
        legend.text = element_text(size = 12),
        axis.title = element_text(size = 15),
        axis.text = element_text(size = 12))
p_legend <- get_legend(p)
pdf(file = paste0(dir_out, "FigS1D.pdf"), width = 7, height = 7)
p + theme(legend.position = "none")
grid.arrange(p_legend)
dev.off()


## S1G: tissue-batch
ami <- AMI(meta$seurat_clusters, meta$batch)
ami_label <- paste0("AMI = ", sprintf("%.3f", ami))
ami_tissue <- AMI(meta$seurat_clusters, meta$tissue)

cat("AMI (batch) =", ami, "\n")
cat("AMI (tissue) =", ami_tissue, "\n")

p <- ggplot(meta, aes(x = UMAP_1, y = UMAP_2, color = batch)) + 
  geom_point(size = 0.1) +
  labs(color = "Batch", title = paste0("All tissues (", ami_label, ")"), x = "", y = "") +
  guides(color = guide_legend(override.aes = list(size = 5))) +
  theme_classic() +
  theme(legend.title = element_text(size = 15),
        legend.text = element_text(size = 12),
        axis.title = element_text(size = 15),
        axis.text = element_text(size = 12))
p_legend <- cowplot::get_legend(p)

plist <- list("all" = p + theme(legend.position = "none"))
for (i in tissue_vec) {
  meta_sub <- meta
  meta_sub$show_point <- meta_sub$tissue == i
  ami <- AMI(meta_sub$seurat_clusters, meta_sub$batch)
  ami_label <- paste0("AMI = ", sprintf("%.3f", ami))
  plist[[i]] <- ggplot(meta_sub, aes(x = UMAP_1, y = UMAP_2, color = batch)) + 
  geom_point(aes(shape = show_point), size = 0.1) +
  scale_shape_manual(values = c("TRUE" = 19, "FALSE" = NA)) +
  labs(color = "Batch", title = i, x = "", y = "") +
  guides(color = guide_legend(override.aes = list(size = 5))) +
  theme_classic() +
  theme(legend.position = "none",
        legend.title = element_text(size = 15),
        legend.text = element_text(size = 12),
        axis.title = element_text(size = 15),
        axis.text = element_text(size = 12))
}

pdf(file = paste0(dir_out, "FigS1G.pdf"), width = 10, height = 6)
grid.arrange(grobs = plist, nrow = 2, right = p_legend, 
             left = textGrob("UMAP_2", gp = gpar(fontsize = 15), rot = 90),
             bottom = textGrob("UMAP_1", gp = gpar(fontsize = 15), rot = 0))
dev.off()
```

## Figure S1E: Heatmap of cluster DE genes (Non-hashing UT GFP+ cells)
```{r}
date_de <- "2020-04-16"
date_cluster <- "2020-03-25"
de_ut_gfppos_cluster <- readRDS(paste0(dir_proj, "2_pipeline/differential_expression/Cluster_UT_GFPpos_inter/", date_de, "/results.rds"))
meta_ut_gfppos <- read.table(paste0(dir_proj, "2_pipeline/clustering/UT_GFPpos_inter/", date_cluster, "/FILES/meta_data.txt"))
so <- so_all[,rownames(meta_ut_gfppos)]
so@meta.data <- meta_ut_gfppos
DefaultAssay(object = so) <- "integrated"
tissue_vec <- c("SPL", "MLN", "PP", "SI", "COL")

## DE genes to plot
thresh_fdr <- 0.05
thresh_log2FC <- log2(1.5)
deg <- lapply(names(de_ut_gfppos_cluster), function(x){
  de_ut_gfppos_cluster[[x]] %>% rownames_to_column(var = "Gene") %>% filter(FDR < thresh_fdr & logFC > thresh_log2FC) %>% mutate(Cluster = x) %>% arrange(-logFC)
}) %>% Reduce(rbind, .)

## heatmap by expression
set.seed(123)
cells_sample <- sample(colnames(so), 1000)
ann_col_df <- data.frame(
  Tissue = factor(so@meta.data[cells_sample, "tissue"], tissue_vec),
  Cluster = factor(so@meta.data[cells_sample, "seurat_clusters"])
)
rownames(ann_col_df) <- cells_sample
ann_col_df <- ann_col_df[order(ann_col_df$Cluster, ann_col_df$Tissue),]
cluster_colors <- ggplot_default_color_descrete(length(unique(so$seurat_clusters)))
names(cluster_colors) <- as.character(0:max(as.numeric(so$seurat_clusters)))
ann_color_ls <- list(
  Cluster = cluster_colors,
  Tissue = tissue_colors(tissue_vec)
)
gaps_col <- cumsum(table(ann_col_df$Cluster))
expr_mat <- so[["integrated"]]@data[intersect(deg$Gene, rownames(so[["integrated"]])), rownames(ann_col_df)]

col <- colorRampPalette(c("magenta", "black", "yellow"))(100)
breaks <- seq(-3, 3, length.out = 101)
gaps_row <- cumsum(table(deg$Cluster))
legend_breaks <- seq(-3, 3, length.out = 7)
legend_labels <- paste0(c("<=", rep("", 5), ">="), legend_breaks)

p <- pheatmap(expr_mat,
         color = col,
         breaks = breaks,
         annotation_col = ann_col_df,
         annotation_colors = ann_color_ls,
         gaps_col = gaps_col,
         legend = T,
         legend_breaks = legend_breaks,
         legend_labels = legend_labels,
         cluster_cols = F,
         cluster_rows = F,
         show_rownames = F,
         show_colnames = F,
         silent = F,
         main = "")

pdf(file = paste0(dir_out, "FigS1E.pdf"), width = 7, height = 7)
grid.arrange(p$gtable)
dev.off()
```

## Figure S1F: Tissue vs. cluster
```{r}
tissue_vec <- c("SPL", "MLN", "PP", "SI", "COL")
plt_df <- meta %>% 
  group_by(seurat_clusters, tissue) %>% tally(name = "n_cluster_tissue") %>% 
  mutate(tissue = factor(tissue, levels = tissue_vec)) %>% arrange(tissue) %>% 
  group_by(seurat_clusters) %>% 
  mutate(n_cluster = sum(n_cluster_tissue), pct = 100 * n_cluster_tissue / n_cluster) %>% 
  mutate(text_pos = n_cluster - (cumsum(n_cluster_tissue) - n_cluster_tissue/2),
         text_lab = ifelse(pct > 10, paste0(sprintf("%.0f", pct), "%"), NA))

p <- ggplot(plt_df) +
  geom_bar(aes(x = factor(seurat_clusters), y = n_cluster_tissue, fill = tissue), stat = "identity") +
  geom_text(aes(x = factor(seurat_clusters), y = text_pos, label = text_lab)) +
  scale_fill_manual(values = tissue_colors()) +
  labs(x = "Cluster", y = "Number of Cells", fill = "Tissue") +
  theme_bw() +
  theme(legend.title = element_text(size = 15, color = "black"),
        legend.text = element_text(size = 12, color = "black"),
        axis.title = element_text(size = 15, color = "black"),
        axis.text = element_text(size = 12, color = "black"))
p
pdf(file = paste0(dir_out, "FigS1F.pdf"), width = 7, height = 7)
p
dev.off()
```

## Figure S1H: distances between tissues
```{r}
library(pheatmap)
## load datas
date_clustering <- "2020-04-13"
meta <- read.table(paste0(dir_proj, "2_pipeline/clustering/UT_GFPpos_inter/", date_clustering, "/FILES/meta_data.txt"),
                   sep = " ")
reductions <- readRDS(paste0(dir_proj, "2_pipeline/clustering/UT_GFPpos_inter/", date_clustering, "/FILES/reductions.rds"))

non_conventional_clusters <- c(6, 7) # tregs, proliferating

meta_sub <- meta[!(meta$seurat_clusters %in% non_conventional_clusters),] %>% droplevels()
pca_sub <- reductions$pca@cell.embeddings[rownames(meta_sub), ]

tissue_vec <- c("SPL", "MLN", "PP", "SI", "COL")

pca_tissue_centers <- sapply(tissue_vec, function(tissue){
  colMeans(pca_sub[meta_sub$tissue == tissue,])
})

pca_dist <- dist(t(pca_tissue_centers)) %>% as.matrix()


col2 <- colorRampPalette(c("#FFFFFF", "#D1E5F0", "#92C5DE",
                           "#4393C3", "#2166AC", "#053061"))
p <- pheatmap(pca_dist,
         color = col2(50),
         cluster_cols = F,
         cluster_rows = F,
         fontsize = 12,
         display_numbers = F,
         silent = T)
p$gtable$grobs[[2]]$gp$col = tissue_colors(tissue_vec)
p$gtable$grobs[[3]]$gp$col = tissue_colors(tissue_vec)
p$gtable$grobs[[2]]$rot = 0
p$gtable$grobs[[2]]$hjust = 0.5
p$gtable$grobs[[2]]$vjust = 1

pdf(file = paste0(dir_out, "FigS1H.pdf"), width = 3.5, height = 3)
grid.arrange(p$gtable)
dev.off()
```


# Figure S2
```{r}
dir_out <- paste0(dir_proj, "3_output/Figure_S2/")
if(!dir.exists(dir_out)) dir.create(dir_out, recursive = T)
```

## Figure S2A: UMAP of UT GFP+/- cells; color by tissue
```{r}
date_cluster <- "2020-03-25"
meta <- read.table(paste0(dir_proj, "2_pipeline/clustering/UT_GFPall_inter/",date_cluster, "/FILES/meta_data.csv"), 
                   header = T, sep = ",", row.names = 1)
coords_umap <- read.table(paste0(dir_proj, "2_pipeline/clustering/UT_GFPall_inter/", date_cluster, "/FILES/coords_umap.csv"),
                          header = T, sep = ",", row.names = 1)
stopifnot(all(rownames(meta) == rownames(coords_umap)))

so <- so_all[,rownames(meta)] ## restrict to SPL EAE cells
so@meta.data <- meta
so$tissue_cluster <- paste(so$tissue, so$seurat_clusters, sep = "_")

so <- NormalizeData(so, assay = "RNA", scale.factor = 10^6) ## compute log(CPM+1)

stopifnot(all(rownames(coords_umap) == rownames(meta)))
meta$UMAP_1 <- coords_umap[,"UMAP_1"]
meta$UMAP_2 <- coords_umap[,"UMAP_2"]

p <- ggplot(meta, aes(x = UMAP_1, y = UMAP_2, color = tissue)) + 
  geom_point(size = 0.1) +
  scale_color_manual(values = tissue_colors()) +
  labs(color = "Tissue") +
  guides(color = guide_legend(override.aes = list(size = 5))) +
  theme_classic() +
  theme(legend.title = element_text(size = 15),
        legend.text = element_text(size = 12),
        axis.title = element_text(size = 15),
        axis.text = element_text(size = 12))
p_legend <- get_legend(p)
pdf(file = paste0(dir_out, "FigS2A.pdf"), width = 7, height = 7)
p + theme(legend.position = "none")
grid.arrange(p_legend)
dev.off()
```


## Figure S2B: UMAP of UT GFP+/- cells; color by expression of Foxp3 and Mki67
```{r}
## Foxp3
meta$log_cpm <- so[["RNA"]]@data["Foxp3",]
p1 <- ggplot(meta, aes(x = UMAP_1, y = UMAP_2, color = log_cpm)) + 
  geom_point(size = 0.1) +
  scale_color_gradient(low = "grey", high = "blue") +
  labs(color = "Expression", title = "Foxp3") +
  theme_classic() +
  theme(legend.title = element_text(size = 12),
        legend.text = element_text(size = 10),
        plot.title = element_text(size = 15, face = "italic"),
        axis.title = element_text(size = 15),
        axis.text = element_blank(), axis.ticks = element_blank())

## Mki67
meta$log_cpm <- so[["RNA"]]@data["Mki67",]
p2 <- ggplot(meta, aes(x = UMAP_1, y = UMAP_2, color = log_cpm)) + 
  geom_point(size = 0.1) +
  scale_color_gradient(low = "grey", high = "blue") +
  labs(color = "Expression", title = "Mki67") +
  theme_classic() +
  theme(legend.title = element_text(size = 12),
        legend.text = element_text(size = 10),
        plot.title = element_text(size = 15, face = "italic"),
        axis.title = element_text(size = 15),
        axis.text = element_blank(), axis.ticks = element_blank())

pdf(file = paste0(dir_out, "FigS2B.pdf"), width = 4.5, height = 7)
grid.arrange(p1, p2, nrow = 2)
dev.off()
```

## Figure S2C: UMAP of UT GFP+/- cells; color by GFP positivity
```{r}
meta$ex_curr <- ifelse(meta$GFP_positive, "Current Th17", "Ex Th17")
p <- ggplot(meta, aes(x = UMAP_1, y = UMAP_2, color = ex_curr)) + 
  geom_point(size = 0.1) +
  scale_color_manual(values = gfp_colors()) +
  labs(color = "Status") +
  guides(color = guide_legend(override.aes = list(size = 5))) +
  theme_classic() +
  theme(legend.title = element_text(size = 15),
        legend.text = element_text(size = 12),
        axis.title = element_text(size = 15),
        axis.text = element_text(size = 12))
p_legend <- get_legend(p)
pdf(file = paste0(dir_out, "FigS2C.pdf"), width = 7, height = 7)
p + theme(legend.position = "none")
grid.arrange(p_legend)
dev.off()
```

## Figure S2D: dot plot of selected Th17 related genes
```{r}
glist <- read.xlsx(paste0(dir_proj, "0_data/gene_lists/200304 Fig1_dotplot.xlsx"), 
                   sheet = "final list", rowNames = F, colNames = F)[,1] %>% rev()
glist <- factor(glist, levels = glist)
tissue_vec <- c('SPL', 'MLN', 'PP', 'SI', 'COL')

## stats for dot plot
ct_mtx = so@assays[['RNA']]@data[glist,] %>% as.matrix()
pct_0 = sapply(tissue_vec, function(x){
  rowMeans(ct_mtx[,so$tissue == x] == 0) * 100
})

exp_mtx = so@assays[['integrated']]@data[glist,] %>% as.matrix()
ave_expr = sapply(tissue_vec, function(x){
  rowMeans(exp_mtx[,so$tissue == x])
}) %>% apply(., 1, scale) %>% t

plt_df <- data.frame(expr = as.numeric(ave_expr), pct = 100 - as.numeric(pct_0), 
           tissue = rep(tissue_vec, each = length(glist)) %>% factor(., levels = rev(tissue_vec)),
           gene = rep(glist, length(tissue_vec)) %>% factor(., levels = rev(glist))) 
p <- ggplot(plt_df) +
  geom_point(aes(x = gene, y = tissue, size = pct, color = expr)) +
  labs(x = "", y = "") +
  scale_colour_gradient(low = "lightgrey", high = "blue") +
  scale_size(range = c(0, 6), trans = scales::trans_new("square", function(x) x^2, "sqrt")) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, face = "italic"), 
        axis.text.y = element_text(color = rev(as.character(tissue_colors(tissue_vec)))),
        legend.title = element_text(size = 8),
        legend.box = "horizontal", legend.position="bottom", legend.key.size = unit(.13, "in")) +
  guides(color = guide_colorbar(title = "Scaled Average\nExpression", title.vjust = .8),
         size = guide_legend(title = "Percent Cell\nDetected", title.vjust = .5, ))


pdf(paste0(dir_out, "FigS2D.pdf"), height = 3, width = 7)
p
dev.off()
```

## Figure S2E: heatmap of fold changes vs SPL
```{r}
date_de <- "2020-04-17"
source(paste0(dir_proj, "1_code/make_figures_tables/utils.R"))
tissue_vec <- c("MLN", "PP", "SI", "COL")

####### plot heatmap ##########
# load differentially expressed genes
dir_de <- paste0(dir_proj, "2_pipeline/differential_expression/UT_GPFall_vs_SPL/", date_de, "/")
deg_up <- readRDS(paste0(dir_de, "up_all.rds"))
deg_down <- readRDS(paste0(dir_de, "down_all.rds"))

# reorder tissue in DE gene lists
deg_up <- deg_up[tissue_vec]
deg_up[sapply(deg_up, is.null)] <- NULL
deg_down <- deg_down[tissue_vec]
deg_down[sapply(deg_down, is.null)] <- NULL

# process gene lists (one list per tissue) into modules (specific to tissue combinations)
combn_to_remove <- c()
up_modules <- gene_lists_to_modules(deg_up, collapse_mLN = F)
down_modules <- gene_lists_to_modules(deg_down, collapse_mLN = F)

all_up_genes <- unlist(up_modules) %>% unique()
down_modules <- lapply(down_modules, function(x) {
  lapply(x, function(y) {setdiff(y, all_up_genes)})
})

# combn_to_remove <- c("mLN_PP_SI", "mLN_PP_colon", "mLN_SI_colon", "mLN_SI", "mLN_colon")
combn_to_remove_up <- sapply(up_modules, function(x) {names(which(sapply(x, length) < 20))}) %>% unlist
up_modules <- lapply(up_modules, function(x){
  x[names(x) %in% combn_to_remove_up] <- NULL
  x
})

combn_to_remove_down <- sapply(down_modules, function(x) {names(which(sapply(x, length) < 20))}) %>% unlist
down_modules <- lapply(down_modules, function(x){
  x[names(x) %in% combn_to_remove_down] <- NULL
  x
})

keep_module_names <- union(unlist(lapply(up_modules, names)), unlist(lapply(down_modules, names)))

# load log fold changes
lfc <- readRDS(paste0(dir_de, "sample_lfc.rds"))
# reorder tissues
samples_ordered <- unlist(lapply(tissue_vec, function(x){which(gsub("_b[0-9]", "", colnames(lfc)) == x)}))
lfc <- lfc[, samples_ordered]
log2fc <- log2(exp(lfc)) # convert to log2 fold change

# reorder genes
genes_orderedrd_up <- lapply(up_modules, function(x){
  lapply(x, function(y){
    if (length(y) == 0) {
      return(NULL)
    } else if (length(y) == 1) {
      return(y)
    } else {
      y[hclust(dist(log2fc[y,]), method = "single")$order]
    }
  })
})  %>% unlist()
genes_orderedrd_down <- lapply(down_modules, function(x){
  lapply(x, function(y){
    if (length(y) == 0) {
      return(NULL)
    } else if (length(y) == 1) {
      return(y)
    } else {
      y[hclust(dist(log2fc[y,]), method = "single")$order]
    }
  })
})  %>% unlist()
genes_orderedrd_down <- genes_orderedrd_down[!genes_orderedrd_down %in% genes_orderedrd_up]
genes_orderedrd <- c(genes_orderedrd_up, genes_orderedrd_down)

de_combn_vec <- sub("[0-9]+", "", names(genes_orderedrd))
de_combn_vec_up <- sub("[0-9]+", "", names(genes_orderedrd_up))
de_combn_vec_down <- sub("[0-9]+", "", names(genes_orderedrd_down))
ng_vec <- c(sapply(unique(de_combn_vec_up), function(x){sum(de_combn_vec_up == x)}),
             sapply(unique(de_combn_vec_down), function(x){sum(de_combn_vec_down == x)})) %>% cumsum

# save genes
for (i in unique(de_combn_vec_up)) {
  write.table(genes_orderedrd_up[de_combn_vec_up == i], file = paste0(dir_de, "module_up_", i, ".txt"),
              sep = "\n", row.names = F, col.names = F, quote = F)
}
for (i in unique(de_combn_vec_down)) {
  write.table(genes_orderedrd_down[de_combn_vec_down == i], file = paste0(dir_de, "module_down_", i, ".txt"),
              sep = "\n", row.names = F, col.names = F, quote = F)
}

# restrict to DE genes; reorder columns
col_order <- c('MLN_b1', 'MLN_b4', 'MLN_b5', 'PP_b1', 'PP_b4', 'PP_b5',
               'SI_b1', 'SI_b3', 'SI_b4', 'COL_b1', 'COL_b2', 'COL_b3', 'COL_b4')
log2fc <- log2fc[genes_orderedrd, col_order]

#### column and row annotations
ann_col <- data.frame(
  tissue = factor(gsub("_b[0-9]", "", colnames(log2fc)), levels = tissue_vec)
)
rownames(ann_col) <- colnames(log2fc)
ann_row <- data.frame(
  DE_in = factor(de_combn_vec, levels = keep_module_names)
)
rownames(ann_row) <- rownames(log2fc)

glist <- c("Gzmb", "Il17f", "Il17a-GFP", "Ifng", "Pdcd1", "Il21", "Klf6", "Nkg7", "Stat1", "Cxcr5", "Sat1", "Cd44",
           "Il7r", "Klf2", "Il17ra", "Tcf7", "Dtx1")
all(glist %in% rownames(log2fc))

library(RColorBrewer)
qual_col_pals = brewer.pal.info[brewer.pal.info$category == 'qual',]
col_vector = unlist(mapply(brewer.pal, qual_col_pals$maxcolors, rownames(qual_col_pals)))
set.seed(12321)
ann_colors = list(
  DE_in = c(sample(col_vector, nlevels(ann_row$DE_in) - nlevels(ann_col$tissue)),
            tissue_colors(tissue_vec)),
  tissue = tissue_colors(tissue_vec)
)
names(ann_colors$tissue) <- levels(ann_col$tissue)
names(ann_colors$DE_in) <- levels(ann_row$DE_in)

lfc_mat_cap <- log2fc
cap_val_high <- 2.5
cap_val_low <- -2.5
lfc_mat_cap[lfc_mat_cap > cap_val_high] <- cap_val_high
lfc_mat_cap[lfc_mat_cap < cap_val_low] <- cap_val_low
my_palette <- c(
  colorRampPalette(c("darkblue", "darkblue"))(1),
  colorRampPalette(c("darkblue", "white", "red"))(100),
  colorRampPalette(c("red", "red"))(1))
tmp <- seq(0, cap_val_high-0.001, length.out = 51)^(0.6)

color_breaks <- c(cap_val_low,
                  rev(-tmp), tmp[-1],
                  cap_val_high)
legend_breaks <- c(seq(cap_val_low, 0, length.out = 3), seq(0, cap_val_high, length.out = 3)) %>% unique()
legend_labels <- paste0(c("<=", rep("", 3), ">="), legend_breaks)

library(pheatmap)
p <- pheatmap(
  lfc_mat_cap,
  cluster_rows = F,
  cluster_cols = F,
  annotation_col = ann_col,
  gaps_row = ng_vec,
  annotation_colors = ann_colors,
  fontsize_col = 6,
  breaks = color_breaks,
  color = my_palette,
  legend_breaks = legend_breaks,
  legend_labels = legend_labels,
  show_rownames = F
)
pdf(file = paste0(dir_out, "FigS2E.pdf"),width = 4)
grid.arrange(p$gtable)
dev.off()
```

## Figure 2F: Differential expression analysis of ex- vs. current- Th17 cells
```{r}
tissue_vec <- c("SPL", "MLN", "PP", "SI", "COL")
res <- sapply(tissue_vec, simplify = F, FUN = function(tissue) {
  readRDS(paste0(dir_proj, "2_pipeline/differential_expression/UT_GFPpos_vs_GFPneg/2021-08-31/", tissue, "_results.rds"))$table
})

thresh_fdr <- 0.05
thresh_lfc <- log2(1.5)
plt_ls <- lapply(res, function(x) {
  x %>% rownames_to_column(var = "gene") %>% 
    mutate(up_in = ifelse(FDR >= thresh_fdr, "ns", ifelse(logFC > thresh_lfc, "GFP+", ifelse(logFC < -thresh_lfc, "GFP-", "ns"))),
           rank_pval = rank(PValue), rank_lfc_inc = rank(logFC), rank_lfc_dec = rank(-abs(logFC)),
           lp = -log10(PValue)) %>% 
    arrange(-abs(logFC))
})

plist <- lapply(tissue_vec, function(x){
  plt_df <- plt_ls[[x]]
  if (sum(plt_df$up_in != "ns") >= 30) {
    plt_df <- plt_df %>% mutate(gene_label = ifelse(up_in != "ns" & (rank_pval < 10 | rank_lfc_inc < 10 | rank_lfc_dec < 10), gene, NA))
  } else {
    plt_df <- plt_df %>% mutate(gene_label = ifelse((up_in != "ns" | rank_pval < 10 | rank_lfc_inc < 10 | rank_lfc_dec < 10), gene, NA))
  }
  if (x == "SPL") {
    plt_df$lp <- pmin(plt_df$lp, 15)
    plt_df$logFC <- pmin(plt_df$logFC, 0.5)
  } else if (x == "MLN") {
    plt_df$lp <- pmin(plt_df$lp, 25)
    plt_df$logFC <- pmin(plt_df$logFC, 1.25)
  } else if (x == "PP") {
    plt_df$lp <- pmin(plt_df$lp, 200)
    plt_df$logFC <- pmin(plt_df$logFC, 2)
  } else if (x == "SI") {
    plt_df$lp <- pmin(plt_df$lp, 100)
    plt_df$logFC <- pmin(plt_df$logFC, 2)
  } else if (x == "COL") {
    plt_df$lp <- pmin(plt_df$lp, 100)
    plt_df$logFC <- pmin(plt_df$logFC, 1.75)
  } else {
    stop("Invalid tissue!!!")
  }
  
  p <- ggplot(plt_df, aes(x = logFC, y = lp, color = up_in, label = gene_label)) +
    geom_point(size = 0.5) +
    geom_text_repel(size = 2, segment.size = 0.1, seed = 1) +
    scale_color_manual(values = c("GFP+" = "darkgreen", "GFP-" = "grey20", "ns" = "grey80")) +
    scale_x_continuous(limits = c(-max(plt_df$logFC), max(plt_df$logFC)), expand = expansion(mult = c(0.01, 0.01))) +
    scale_y_continuous(limits = c(0, max(plt_df$lp)), expand = expansion(mult = c(0.05, 0.01))) +
    # labs(title = x, caption = paste0("FDR < ", thresh_fdr, ", fold change > ", sprintf("%.2f", 2^thresh_lfc))) +
    labs(title = x, color = "Up-regulated in", y = "-log10(p)", x = "Log2 fold change (GFP+ vs. GFP-)") +
    theme_bw() +
    theme(plot.title = element_text(size = 12, hjust = 0.5, face = "bold"),
          axis.title = element_text(size = 10, color = "black"))
  return(p)
})

pdf(file = paste0(dir_out, "FigS2F.pdf"), width = 10, height = 6)
grid.arrange(grobs = c(lapply(plist, function(x){x + theme(legend.position = "none")}), list(get_legend(plist[[3]]))), nrow = 2)
dev.off()
```


# Figure S3
```{r}
dir_out <- paste0(dir_proj, "3_output/Figure_S3/")
if(!dir.exists(dir_out)) dir.create(dir_out, recursive = T)
```

## Figure S3D: Cd274 in other data
```{r}
library(epitools)
## load data
data_names <- c("Tibbitt", "Hemmers", "Arazi")
df_cd274 <- sapply(data_names, simplify = FALSE, FUN = function(i) {
  readRDS(paste0(dir_proj, "2_pipeline/Pdl1_", i, "/df_cd274_2020-07-23.rds"))
})

## Odds ratios
OR <- sapply(data_names, function(x) {
  # RRtable <- table(df_cd274[[x]]$expr > 0, df_cd274[[x]]$in_cluster) 
  # riskratio.wald(RRtable)
  
  ORtable <- table(df_cd274[[x]]$expr > 0, df_cd274[[x]]$in_cluster) 
  OR_res <- oddsratio.wald(ORtable)
  
  c(OR_res$measure[2,], OR_res$p.value[2,])
})

plt_df <- data.frame(t(OR)) %>% rownames_to_column(var = "Data") %>% 
  mutate(log_estimate = log(estimate),
         log_lower = log(lower),
         log_upper = log(upper)) %>% 
  mutate(Data = factor(Data, levels = data_names))
  
p_OR <- ggplot(plt_df) +
  geom_point(aes(x = Data, y = estimate), size = 2) +
  geom_errorbar(aes(x = Data, ymin = lower, ymax = upper), width = 0.2) +
  geom_hline(yintercept = 1, linetype = 2) +
  ylim(0, max(plt_df$upper)) +
  labs(x = "", y = "Odds Ratio") +
  coord_flip() +
  theme_bw() +
  theme(legend.position = "none",
        axis.text.y = element_text(size = 12, color = "black"))


pdf(paste0(dir_out, "FigS3D.pdf"), width = 4, height = 2)
p_OR
dev.off()
```

## Figure S3E: clonal similarity between ISG-high clusters and all others
```{r}
analysis_date <- "2020-11-02"
sim_scores <- readRDS(paste0(dir_proj, "2_pipeline/TCR/compare_tissues/similarity_score_tissue_cluster_UT_GFPall_", analysis_date, ".rds"))

isg_high <- c("SPL_5", "MLN_4", "PP_5")

tissue_vec <- c("SPL", "MLN", "PP", "SI", "COL")
all_clusters <- lapply(tissue_vec, function(x){
  colnames(sim_scores)[grep(x, colnames(sim_scores))] %>% sort()
}) %>% unlist()

diag(sim_scores) <- NA

plt_df <- sim_scores[isg_high,] %>% data.frame() %>% rownames_to_column(var = "cluster1") %>% 
    gather(key = "cluster2", value = "similarity", -cluster1) %>% 
    mutate(cluster2 = factor(cluster2, levels = all_clusters),
           cluster1 = factor(cluster1, levels = rev(isg_high)))
  
p <- ggplot(plt_df) +
  geom_point(aes(x = cluster2, y = cluster1, fill = similarity, size = similarity), shape = 21) +
  geom_segment(data = data.frame(xend = seq(0.5, ncol(sim_scores) + 1)),
               aes(x = xend, xend = xend, y = 0.5, yend = length(isg_high) + 0.5), color = "grey85") +
  geom_hline(yintercept = seq(0.5, length(isg_high) + 1, 1), color = "grey85") +
  scale_size_continuous(name = "Similarity", range = c(0, 4), 
                        breaks = seq(0, 0.3, 0.1), labels = c("0", "0.1", "0.2", ">=0.3")) +
  scale_fill_gradientn(name = "Similarity", 
                       colours = rev(c("#67001F", "#B2182B", "#D6604D", "#F4A582", "#FDDBC7", "#FFFFFF")),
                       breaks = seq(0, 0.3, 0.1), labels = c("0", "0.1", "0.2", ">=0.3")) +
  guides(fill = guide_legend(), size = guide_legend()) +
  scale_x_discrete(position = "top") +
  labs(x = "", y = "", title = "") + 
  coord_fixed(ratio = 1) +
  theme_classic() +
  theme(axis.ticks=element_blank(),
        axis.text.x=element_text(size = 10, angle = 90, hjust = 0,
                                 colour = tissue_colors(gsub("_.*", "",all_clusters))),
        axis.text.y=element_text(size = 10, angle = 0, hjust = 0,
                                   colour = rev(tissue_colors(gsub("_.*", "", isg_high)))),
        axis.line=element_blank(),
        legend.position = "bottom",
        plot.margin = unit(c(10,10,10,10), "mm"))
p
pdf(file = paste0(dir_out, "FigS3E.pdf"), width = 7, height = 7)
p
dev.off()
```


## Figure S3H: clonal similarity between clusters within tissue
```{r}
clustering_date <- "2020-03-25"
dir_clustering <- paste0(dir_proj, "2_pipeline/clustering/UT_GFPall_intra/", clustering_date, "/")

tissue_vec <- c("SPL", "MLN", "PP", "SI", "COL")
### load clustering data
meta <- list()
reductions <- list()
for (tissue in tissue_vec) {
  meta[[tissue]]  <- read.table(paste0(dir_clustering, "FILES/", tissue, "/meta_data.csv"), header = T, row.names = 1, sep = ",")
  meta[[tissue]] <- meta[[tissue]][,-grep("integrated_snn_res", colnames(meta[[tissue]]))]
  meta[[tissue]]$tissue_cluster <- paste(meta[[tissue]]$tissue, meta[[tissue]]$seurat_clusters, sep = "_")
  reductions[[tissue]] <- readRDS(paste0(dir_clustering, "FILES/", tissue, "/reductions.rds"))
}
all_meta <- Reduce(rbind, meta)
tissue_cluster_vec <- unique(all_meta$tissue_cluster)

## load annotation
cluster_ann_types <- c("Proliferating", "Treg-like", "Effector-like", "Tfh-like", "ISG-high", "Cxcr6+", "Other")
cluster_ann <- read.xlsx(paste0(dir_clustering, "cluster_annotation.xlsx"), sheet = "processed")
cluster_ann <- cluster_ann[!duplicated(cluster_ann),]
cluster_ann$Annotation[is.na(cluster_ann$Annotation)] <- "Other"
stopifnot(length(setdiff(cluster_ann$Cluster, tissue_cluster_vec)) == 0) ## check if there are wierd cluster in annotation file
if (length(setdiff(tissue_cluster_vec, cluster_ann$Cluster)) > 0) {
  other_clusters <- data.frame(
    Cluster = setdiff(tissue_cluster_vec, cluster_ann$Cluster),
    Annotation = "Other")
  cluster_ann <- rbind(cluster_ann, other_clusters)
}
cluster_ann$Cluster <- new_cluster_vec
## set levels
cluster_ann$Annotation <- factor(cluster_ann$Annotation, levels = cluster_ann_types)
cluster_ann <- cluster_ann[order(cluster_ann$Annotation),]

## cluster colors
max_n_cluster <- sapply(strsplit(tissue_cluster_vec, split = "_"), function(x){x[2]}) %>% as.numeric() %>% max() + 1
color_cluster <- brewer.pal(max_n_cluster, "Set1")
names(color_cluster) <- 0:(max_n_cluster - 1)
## cell type colors
color_ann <- brewer.pal(nlevels(cluster_ann$Annotation), "Dark2")
color_ann[nlevels(cluster_ann$Annotation)] <- "grey50"
names(color_ann) <- levels(cluster_ann$Annotation)
## clonotypes
clonotyping_type_date <- "dominant_2020-04-02"
clty_all <- read.csv(paste0(dir_proj, "2_pipeline/TCR/clonotype_assignment_", clonotyping_type_date, ".csv"), row.names = 1)

## merge clonotype info to meta data
all_meta$clonotype_id <- clty_all[rownames(all_meta),"clonotype_id"]
all_meta$pooled_clonotype_id <- paste(all_meta$treatment, all_meta$batch, all_meta$mouse, all_meta$clonotype_id, sep = "_")
all_meta$tissue_cluster <- paste(all_meta$tissue, all_meta$seurat_clusters, sep = "_")
all_meta$sample <- paste(all_meta$batch, all_meta$mouse, sep = "_") %>% factor()
meta_noNA <- all_meta[!is.na(all_meta$clonotype_id),] %>% droplevels()

stopifnot(all(meta_noNA$treatment == "UT")) ## check if treatment is correct (should be UT cells only)

cluster_colors <- color_ann[cluster_ann$Annotation]
names(cluster_colors) <- cluster_ann$Cluster

## compute similarity
pdf(paste0(dir_out, "FigS3H.pdf"), width = 5.5, height = 5)
for (tissue in c("MLN", "PP", "SI")) {
  cl_mtx <- table(as.character(meta_noNA$pooled_clonotype_id[meta_noNA$tissue == tissue]), 
                meta_noNA$tissue_cluster[meta_noNA$tissue == tissue]) %>% as.matrix()
  cl_mtx <- apply(cl_mtx, 2, function(x){x/sum(x)})
  bh_dist <- philentropy::distance(t(cl_mtx), method = "bhattacharyya") %>% as.matrix()
  rownames(bh_dist) <- colnames(cl_mtx)
  colnames(bh_dist) <- colnames(cl_mtx)
  bh_sim <- exp(-bh_dist)
  
  bh_sim_plt <- bh_sim
  diag(bh_sim_plt) <- NA
  corrplot(bh_sim_plt, 
           is.corr = F,
           type="upper", order="original", method = "circle",
           # col=rev(brewer.pal(n=8, name="RdYlBu")),
           col = rev(col_corr_no_white(200)),
           outline = T,
           tl.col = cluster_colors[rownames(bh_sim_plt)],
           tl.pos = 'td',
           tl.cex = 1,
           mar = c(1,1,1,1),
           diag = T, na.label = " ",
           title = tissue)

}
  dev.off()

```


## Figure S3I: TCR sharing with SPL
See Fig2.Rmd


# Figure S4
```{r}
dir_out <- paste0(dir_proj, "3_output/Figure_S4/")
if(!dir.exists(dir_out)) dir.create(dir_out, recursive = T)
```

## Figure S4A: UMAP of G1/S and G2/M signatures in CNS
```{r}
date_cluster <- "2020-03-25"
meta_cns <- read.table(paste0(dir_proj, "2_pipeline/clustering/EAE_GFPall_intra/", date_cluster, "/FILES/CNS/meta_data.csv"),
                       header = T, row.names = 1, sep = ",")
reductions_cns <- readRDS(paste0(dir_proj, "2_pipeline/clustering/EAE_GFPall_intra/", date_cluster, "/FILES/CNS/reductions.rds"))
stopifnot(all(rownames(meta_cns) == rownames(reductions_cns$umap@cell.embeddings)))
so <- so_all[,rownames(meta_cns)]
so@meta.data <- meta_cns
so@reductions <- reductions_cns

s.genes <- human_to_mouse(cc.genes.updated.2019$s.genes)
g2m.genes <- c(human_to_mouse(cc.genes.updated.2019$g2m.genes), "Anp32e")
so <- CellCycleScoring(so, s.features = s.genes, g2m.features = g2m.genes, set.ident = TRUE)

## plot signature
plist <- list()
for (i in c("S.Score", "G2M.Score")) {
  plt_df <- data.frame(so@reductions$umap@cell.embeddings)
  plt_df$score <- so@meta.data[,i]
  max_cutoff <- quantile(plt_df$score, 0.95, na.rm = T)
  min_cutoff <- quantile(plt_df$score, 0.05, na.rm = T)
  plt_df$score_cap <- plt_df$score %>% pmin(., max_cutoff) %>% pmax(., min_cutoff)
  
  title_label <- ifelse(i == "S.Score", "G1/S Signature", "G2/M Signature")
  plist[[i]] <- ggplot(plt_df) +
    geom_point(aes(UMAP_1, UMAP_2, color = score_cap), size = 0.5) +
    scale_color_gradient(low = "grey80", high = "blue", na.value = "grey50") +
    labs(color = "Signature\nscore", title = title_label) +
    theme_classic() +
    theme(plot.title = element_text(face = "bold", hjust = 0, colour = "black"),
          axis.text = element_blank(), axis.ticks = element_blank())
}
pdf(file = paste0(dir_out, "FigS4A.pdf"), width = 9, height = 4)
grid.arrange(grobs = plist, nrow = 1)
dev.off()
```



## Figure S4B: boxplot of total UMIs.
```{r}
meta_cns$tissue_cluster <- paste0("C", meta_cns$seurat_clusters)
p <- ggplot(meta_cns, aes(x = tissue_cluster, y = nFeature_RNA, fill = tissue_cluster)) +
  geom_boxplot(width = 0.7) +
  geom_signif(comparisons = list(c("C4", "C0"), c("C4", "C1"), c("C4", "C3")), 
              test = "t.test", map_signif_level = TRUE, y_position = c(6900, 6450, 6000), textsize = 10, vjust = 0.6) +
  labs(x = "", y = "Total UMIs") +
  ylim(0, 7000) +
  theme_bw() +
  theme(legend.position = "none",
        axis.title.y = element_text(size = 15, color = "black"),
        axis.text.x = element_text(size = 15, color = "black"),
        axis.text.y = element_text(size = 12, color = "black"))
p
pdf(paste0(dir_out, "FigS4B.pdf"), height = 4, width = 4)
p
dev.off()
```

## Figure S4C: proportion of GFP+ cells
```{r}
df_plt <- meta_cns %>% group_by(seurat_clusters) %>% 
  summarise(gfp_pos = 100 * mean(GFP_positive), gfp_neg = 100 - gfp_pos) %>%
  mutate(cluster = paste0("C", seurat_clusters)) %>% 
  select(-seurat_clusters) %>% 
  gather(key = "gfp", value = "pct", -cluster) %>% 
  mutate(status = ifelse(gfp == "gfp_pos", "Current Th17", "Ex Th17"))

p <- ggplot(df_plt) +
  geom_bar(aes(x = cluster, y = pct, fill = status), stat = "identity", width = 0.7, color = "black") +
  scale_fill_manual(values = gfp_colors()) +
  labs(x = "", y = "Percent of Cells", fill = "Status") +
  scale_y_continuous(expand = c(0, 0)) +
  theme_classic() +
  theme(
        axis.text.x = element_text(face = "bold"),
        axis.title.y = element_text(face = "bold"),
        axis.text.y = element_text(face = "bold"))
pdf(paste0(dir_out, "FigS4C.pdf"), width = 3.5, height = 3.5)
p
dev.off()
```




## Figure S4E: UMAP of all cells, color by batch
```{r}
clustering_date <- "2020-04-16"
meta <- read.table(paste0(dir_proj, "2_pipeline/clustering/all_inter/", clustering_date, "/FILES/meta_data.csv"), 
                   header = T, sep = ",", row.names = 1)
reductions <- readRDS(paste0(dir_proj, "2_pipeline/clustering/all_inter/", clustering_date, "/FILES/reductions.rds"))
stopifnot(all(rownames(meta) == rownames(reductions$umap@cell.embeddings)))

plt_df <- data.frame(
  batch = meta$batch,
  UMAP_1 = reductions$umap@cell.embeddings[,"UMAP_1"],
  UMAP_2 = reductions$umap@cell.embeddings[,"UMAP_2"]
)
p <- ggplot(plt_df, aes(x = UMAP_1, y = UMAP_2, color = batch)) + 
  geom_point(size = 0.1) +
  labs(color = "Batch") +
  guides(color = guide_legend(override.aes = list(size = 5))) +
  theme_classic() +
  theme(legend.title = element_text(size = 15),
        legend.text = element_text(size = 12),
        axis.title = element_text(size = 15),
        axis.text = element_text(size = 12))
p_legend <- get_legend(p)
pdf(file = paste0(dir_out, "FigS4E.pdf"), width = 7, height = 7)
p + theme(legend.position = "none")
grid.arrange(p_legend)
dev.off()
```


## Figure S4F: UMAP of all cells, color by GFP+/-
```{r}
plt_df <- data.frame(
  GFP = factor(meta$GFP_positive, levels = c("TRUE", "FALSE"), labels = c("Current Th17", "Ex Th17")),
  UMAP_1 = reductions$umap@cell.embeddings[,"UMAP_1"],
  UMAP_2 = reductions$umap@cell.embeddings[,"UMAP_2"]
)

p <- ggplot(plt_df, aes(x = UMAP_1, y = UMAP_2, color = GFP)) + 
  geom_point(size = 0.1) +
  scale_color_manual(values = gfp_colors()) +
  labs(color = "Status") +
  guides(color = guide_legend(override.aes = list(size = 5))) +
  theme_classic() +
  theme(legend.title = element_text(size = 15),
        legend.text = element_text(size = 12),
        axis.title = element_text(size = 15),
        axis.text = element_text(size = 12))
p_legend <- get_legend(p)
pdf(file = paste0(dir_out, "FigS4F.pdf"), width = 7, height = 7)
p + theme(legend.position = "none")
grid.arrange(p_legend)
dev.off()
```


## Figure S4G: compare distances between EAE and UT for SPL, COL and SI. (in PC space, similar to S3)
```{r}
## restict to conventional Th17 cells only
non_conventional_clusters <- c(7, 8, 10) # tregs, proliferating
meta_sub <- meta[!(meta$seurat_clusters %in% non_conventional_clusters),] %>% droplevels()
pca_sub <- reductions$pca@cell.embeddings[rownames(meta_sub), ]

tissue_vec <- c("SPL", "MLN", "PP", "SI", "COL")
tissue_vec_sub <- c("SPL", "SI", "COL")

pc_dist <- sapply(tissue_vec, function(tissue){
  pc_center_ut <- colMeans(pca_sub[(meta_sub$tissue == tissue) & (meta_sub$treatment == "UT"),])
  pc_center_eae <- colMeans(pca_sub[(meta_sub$tissue == tissue) & (meta_sub$treatment == "EAE"),])
  sqrt(sum((pc_center_ut - pc_center_eae)^2))
})

plt_df <- data.frame(
  tissue = factor(tissue_vec, levels = rev(tissue_vec)),
  distance = pc_dist
)

p <- plt_df %>% filter(tissue %in% tissue_vec_sub) %>% 
  ggplot(aes(x = tissue, y = distance, fill = tissue)) +
  geom_bar(stat = "identity", width = 0.5, color = "black") +
  scale_fill_manual(values = tissue_colors()) +
  labs(x = "", y = "Distance") +
  scale_y_continuous(limits = c(0, 8), expand = c(0, 0)) +
  coord_flip() +
  theme_bw() +
  theme(legend.position = "none",
        axis.title = element_text(size = 15, color = "black"),
        axis.text = element_text(size = 12, color = "black"))

pdf(file = paste0(dir_out, "FigS4G.pdf"), width = 5, height = 3)
p
dev.off()
```


## Figure S4H: compare number of DEG in SPL, COL and SI.
```{r}
date_analysis <- "2020-12-13"
tissue_vec <- c("SPL", "SI", "COL", "MLN", "PP")
de_res <- lapply(tissue_vec, function (tissue) {
  readRDS(paste0(dir_proj, "2_pipeline/differential_expression/EAE_vs_UT_GFPall/", date_analysis, "/", tissue, "_results.rds"))
})
names(de_res) <- tissue_vec

thresh_fdr <- 0.05
thresh_log2FC <- log2(1.5)

de_up <- lapply(names(de_res), function(i){
  x <- de_res[[i]]
  data.frame(gene = rownames(x$table)[x$table$logFC > thresh_log2FC & x$table$FDR < thresh_fdr],
             tissue = i, up_in = "EAE")
})
de_up <- Reduce(rbind, de_up)
de_down <- lapply(names(de_res), function(i){
  x <- de_res[[i]]
  data.frame(gene = rownames(x$table)[x$table$logFC < -thresh_log2FC & x$table$FDR < thresh_fdr],
             tissue = i, up_in = "Naïve")
})
de_down <- Reduce(rbind, de_down)

col_treatment <- ggplot_default_color_descrete(2)
names(col_treatment) <- c("EAE", "Naïve")
col_treatment <- c(col_treatment, "N.S." = "#000000")

tissue_vec_sub <- c("SPL", "COL", "SI")

df_plot <- rbind(de_up, de_down) %>% group_by(up_in, tissue) %>% tally() %>% ungroup() %>% 
  filter(tissue %in% tissue_vec_sub) %>% 
  mutate(tissue = factor(tissue, levels = tissue_vec_sub))
p <- ggplot(df_plot) +
  geom_bar(aes(x = up_in, y = n, fill = tissue), stat = "identity", position = "dodge", width = 0.5, color = "black") +
  geom_hline(yintercept = 0, color = "black") +
  scale_fill_manual(values = tissue_colors(tissue_vec_sub)) +
  labs(x = "Up-regulated Condition", y = "Number of Genes") +
  coord_flip() +
  facet_wrap(~tissue, nrow = 3) +
  theme_bw() +
  theme(legend.position = "none",
        axis.text = element_text(size = 12, color = "black"),
        axis.title = element_text(size = 15, color = "black"),
        strip.text = element_text(size = 15, color = "black"))
p
pdf(file = paste0(dir_out, "FigS4H.pdf"), width = 4, height = 4)
p
dev.off()
```

## Figure S4I: compare SPL, SI and COL
```{r}
tissue_vec_sub <- c("SPL", "SI", "COL")
date_cluster_umap <- "2020-05-07"
date_cluster_meta <- "2020-04-28"
meta_ls <- list()
reduction_ls <- list()
for (tissue in tissue_vec_sub) {
 meta_ls[[tissue]] <- read.table(paste0(dir_proj, "2_pipeline/clustering/all_intra/", 
                                         date_cluster_meta, "/FILES/", tissue, "/meta_data.csv"),
                                  sep = ",", row.names = 1, header = T)
 reduction_ls[[tissue]] <- readRDS(paste0(dir_proj, "2_pipeline/clustering/all_intra/", 
                                         date_cluster_umap, "/FILES/", tissue, "/reductions.rds"))
}

all_meta <- Reduce(rbind, meta_ls)
all_meta$tissue_cluster <- paste(all_meta$tissue, all_meta$seurat_clusters, sep = "_")
tissue_cluster_vec <- unique(all_meta$tissue_cluster)

## cluster colors
max_n_cluster <- sapply(strsplit(tissue_cluster_vec, split = "_"), function(x){x[2]}) %>% as.numeric() %>% max() + 1
max_n_set1_color <- 9
if (max_n_cluster <= max_n_set1_color) {
  color_cluster <- brewer.pal(max_n_cluster, "Set1")
} else {
  color_cluster <- c(brewer.pal(max_n_set1_color, "Set1"), brewer.pal(max_n_cluster - max_n_set1_color, "Set3"))
}
names(color_cluster) <- 0:(max_n_cluster - 1)
## treatment colors
col_treatment <- ggplot_default_color_descrete(2)
names(col_treatment) <- c("EAE", "Naïve")

p_cluster <- list()
p_gfp <- list()
p_trt <- list()
p_bar <- list()

n_ds <- 5 ## number of down samplings to perform
n_ds_cells <- table(all_meta$treatment, all_meta$tissue) %>% as.numeric() %>% min() ## number of cells to sample
set.seed(123)
for (i in 1:n_ds) {
  for (tissue in tissue_vec_sub) {
    cat(tissue, "\n")
    plt_df <- data.frame(
      cluster = factor(meta_ls[[tissue]]$seurat_clusters),
      ex_curr = factor(ifelse(meta_ls[[tissue]]$GFP_positive, "Current Th17", "Ex Th17")),
      treatment = factor(ifelse(meta_ls[[tissue]]$treatment == "UT", "Naïve", "EAE"), levels = c("Naïve", "EAE")),
      UMAP_1 = reduction_ls[[tissue]]$umap@cell.embeddings[,"UMAP_1"],
      UMAP_2 = reduction_ls[[tissue]]$umap@cell.embeddings[,"UMAP_2"]
    )
    plt_df <- plt_df[sample(1:nrow(plt_df), nrow(plt_df), replace = F),]# shuffle cells so UT cells don't always cover EAE cells
    
    ## downsample 
    idx_eae <- sample(which(plt_df$treatment == "EAE"), n_ds_cells, replace = F)
    idx_naive <- sample(which(plt_df$treatment == "Naïve"), n_ds_cells, replace = F)
    plt_df_ds <- plt_df[c(idx_eae, idx_naive),]
    
    p_cluster[[tissue]] <- ggplot(plt_df, aes(x = UMAP_1, y = UMAP_2, color = cluster)) + 
      geom_point(size = 0.1) +
      scale_color_manual(values = color_cluster) +
      guides(color = guide_legend(override.aes = list(size = 5))) +
      labs(x = "UMAP_1", y = "UMAP_2", color = "Cluster", title = tissue) +
      theme_classic() +
      theme(plot.title = element_text(colour = tissue_colors(tissue), hjust = 0, face = "bold", size = 20),
            legend.position = "none",
            axis.ticks = element_blank(), axis.text = element_blank())
    p_gfp[[tissue]] <- ggplot(plt_df_ds, aes(x = UMAP_1, y = UMAP_2, color = ex_curr)) + 
      geom_point(size = 0.1) +
      scale_color_manual(values = gfp_colors()) +
      guides(color = guide_legend(override.aes = list(size = 5))) +
      labs(x = "UMAP_1", y = "UMAP_2", color = "Status") +
      theme_classic() +
      theme(legend.position = "none",
            axis.ticks = element_blank(), axis.text = element_blank())
    
    p_trt[[tissue]] <- ggplot(plt_df_ds, aes(x = UMAP_1, y = UMAP_2, color = treatment)) + 
      geom_point(size = 0.1) +
      scale_color_manual(values = col_treatment) +
      guides(color = col_treatment) +
      labs(x = "UMAP_1", y = "UMAP_2", color = "Treatment") +
      theme_classic() +
      theme(legend.position = "none",
            axis.ticks = element_blank(), axis.text = element_blank())
    
    plt_df_2 <- plt_df_ds %>% 
      group_by(cluster, treatment) %>% tally() %>% 
      group_by(cluster) %>% mutate(total = sum(n))
    
    chisq_out <- table(plt_df_ds$treatment, plt_df_ds$cluster) %>% chisq.test()
    chisq_label <- paste0("Chi-square = ", sprintf("%.2f", chisq_out$statistic))
    chisq_x <- nlevels(plt_df_2$cluster)
    chisq_y <- max(plt_df_2$total)
  
    p_bar[[tissue]] <- ggplot(plt_df_2, aes(x = factor(cluster), y = n, fill = treatment)) +
      geom_bar(width = 0.5, stat = "identity", color = "black", size = 1) +
      scale_fill_manual(values = col_treatment) +
      annotate(geom = "text", x = chisq_x, y = chisq_y, label = chisq_label, size = 5, hjust = 1, vjust = 1) +
      scale_y_continuous(expand = c(0, 0)) +
      labs(x = "Clusters", y = "Number of Cells") +
      theme_classic() +
      theme(legend.position = "none")
  }
  
  ## create legends
  color_cluster_pad <- color_cluster
  names(color_cluster_pad) <- names(color_cluster)
  tmp <- data.frame(x = factor(0:(max_n_cluster-1)), y = 1) %>% 
    ggplot(aes(x, y, color = x)) + 
    scale_color_manual(values = color_cluster_pad) + 
    geom_point() + labs(colour = "Cluster") + 
    theme(legend.text = element_text(size = 15),
          legend.title = element_text(size = 20)) +
    guides(color = guide_legend(override.aes = list(size = 5)))
  p_cluster$legend <- as_ggplot(get_legend(tmp)) 
  
  tmp <- data.frame(x = c("Current Th17", "Ex Th17"), y = 1) %>% 
    ggplot(aes(x, y, color = x)) + 
    scale_color_manual(values = gfp_colors()) + 
    geom_point() + theme_bw() + labs(colour = "Status")+ 
    theme(legend.text = element_text(size = 15),
          legend.title = element_text(size = 20)) +
    guides(color = guide_legend(override.aes = list(size = 5)))
  p_gfp$legend <- as_ggplot(get_legend(tmp))
  
  tmp <- data.frame(x = c("Naïve", "EAE"), y = 1) %>% 
    ggplot(aes(x, y, color = x)) + 
    scale_color_manual(values = col_treatment) + 
    geom_point() + theme_bw() + labs(colour = "Treatment")+ 
    theme(legend.text = element_text(size = 15),
          legend.title = element_text(size = 20)) +
    guides(color = guide_legend(override.aes = list(size = 5)))
  p_trt$legend <- as_ggplot(get_legend(tmp))
  p_bar$legend <- p_trt$legend
  
  
  pdf(file = paste0(dir_out, "FigS4I_", i, ".pdf"), width = 28, height = 8)
  grid.arrange(grobs = c(p_cluster[c(tissue_vec_sub, "legend")], 
                         p_trt[c(tissue_vec_sub, "legend")],
                         p_gfp[c(tissue_vec_sub, "legend")],  
                         p_bar[c(tissue_vec_sub, "legend")]), nrow = 2, ncol = 8
               )
  dev.off()
}
```


# Figure S5
```{r}
dir_out <- paste0(dir_proj, "3_output/Figure_S5/")
if(!dir.exists(dir_out)) dir.create(dir_out, recursive = T)
```

## Figure S5A, B, C: correlation between bulk SLAMF6+ and CXCR6+ signatures vs. known Th17 signatures in human data
```{r}
date_analysis <- "2020-07-01"
meta_pbmc <- readRDS(file = paste0(dir_proj, "2_pipeline/human_shafflick/", date_analysis, "/meta_pbmc.rds"))
meta_csf <- readRDS(file = paste0(dir_proj, "2_pipeline/human_shafflick/", date_analysis, "/meta_csf.rds"))
sig_ls <- readRDS(file = paste0(dir_proj, "2_pipeline/human_shafflick/", date_analysis, "/signature_list.rds"))

#### Correlation between Cxcr6 and Th17 pathogenicity
sig_to_plot <- c("Th17_Th1_like_memory-plus" = "Th17/Th1-like memory",
                 "Pathogenic_Th17" = "Th17 Pathogenic",
                 "Th17_Th1_like_effector_CNS-plus" = "Th17/Th1-like effector CNS",
                 "Th17_Th1_like_effector-plus" = "Th17/Th1-like effector",
                 "Th17_self_renewing-plus" = "Th17 self-renewing")

plist <- list()
for (i in seq_along(sig_to_plot)) {
  sig <- names(sig_to_plot)[i]
  plt_df <- data.frame(
    x = meta_csf$`Cxcr6_bulk-plus`, 
    y = meta_csf[,sig],
    is_MS = meta_csf$is_MS
  ) 
  plt_df <- plt_df %>% filter(is_MS == "MS") %>% 
    mutate(x_z = scale(x), y_z = scale(y)) %>% 
    filter(abs(x_z) < 5, abs(y_z) < 5)
  pearson_corr <- cor.test(plt_df$x_z, plt_df$y_z, alternative = "two.sided", method = "pearson")
  plist[[sig]] <- ggplot(plt_df, aes(x = x_z, y = y_z)) +
    geom_point(alpha = 0.3) +
    geom_smooth(color = "blue", method = 'lm') +
    labs(x = "", y = "",
         title = sig_to_plot[i],
         subtitle = paste0("Spearman-corr=", sprintf("%.3f", pearson_corr$estimate),
                           map_pval(pearson_corr$p.value))) +
    theme_bw() +
    theme(plot.title = element_text(face = "bold", color = "black", size = 12, hjust = 0.5),
          plot.subtitle = element_text(color = "black", size = 12, hjust = 1),
          legend.position = "none")
}

pdf(paste0(dir_out, "FigS5A.pdf"), width = 7, height = 5)
grid.arrange(grobs = plist[setdiff(names(sig_to_plot), "Th17_self_renewing-plus")],
             nrow = 2, ncol = 2,
             left = textGrob(label = "Signature Score", rot = 90, gp = gpar(fontsize = 12)),
             bottom = textGrob(label = "CXCR6+ Signature Score", rot = 0, gp = gpar(fontsize = 12)))
dev.off()


pdf(paste0(dir_out, "FigS5B.pdf"), width = 3.5, height = 2.5)
grid.arrange(plist$`Th17_self_renewing-plus`,
             left = textGrob(label = "Signature Score", rot = 90, gp = gpar(fontsize = 12)),
             bottom = textGrob(label = "CXCR6+ Signature Score", rot = 0, gp = gpar(fontsize = 12)))
dev.off()


corr_lim <- 0.3
corr_breaks <- seq(-corr_lim, corr_lim, length.out = 5)
corr_labels <- paste0(c("\u2264 ", "", "", "", "\u2265 "), corr_breaks)
tmp <- sapply(names(sig_to_plot), function(x) {
  cor_ms <- cor.test(meta_csf[meta_csf$is_MS == "MS",x], meta_csf[meta_csf$is_MS == "MS", "Cxcr6_bulk-plus"], 
                alternative = "two.sided", method = "pearson")
  cor_healthy <- cor.test(meta_csf[meta_csf$is_MS == "Healthy",x], meta_csf[meta_csf$is_MS == "Healthy", "Cxcr6_bulk-plus"], 
                alternative = "two.sided", method = "pearson")
  c(cor_ms$estimate, cor_ms$p.value, cor_healthy$estimate, cor_healthy$p.value)
}) %>% t
plt_df <- rbind(tmp[,1:2], tmp[,3:4]) %>% data.frame()
colnames(plt_df) <- c("corr", "pval")
plt_df$condition <- rep(c("MS", "Healthy"), each = nrow(tmp))
plt_df$signature <- rep(sig_to_plot, times = 2) %>% factor(., levels = rev(sig_to_plot))
plt_df$corr_cap <- pmax(pmin(plt_df$corr, corr_lim), -corr_lim)
p <- ggplot(plt_df, aes(x = condition, y = signature)) +
  geom_point(aes(size = -log10(pval), fill = corr_cap), pch = 21, color = "black") +
  scale_size(range = c(1, 10), breaks = c(100, 200, 300), labels = c("100", "200", "300"), limits = c(0, 300)) +
  scale_fill_gradientn(colours = rev(c("#67001F", "#B2182B", "#D6604D", "#F4A582",
                                    "#FFFFFF", "#92C5DE",
                                    "#4393C3", "#2166AC", "#053061")),
                       limits = c(-corr_lim, corr_lim),
                       breaks = corr_breaks,
                       labels = corr_labels
                       ) +
  labs(x = "", y = "", fill = "Pearson Corr.") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 12, color = "black"),
        axis.text.y = element_text(size = 12, color = "black"))
table(meta_csf$is_MS)

cairo_pdf(paste0(dir_out, "FigS5C.pdf"), width = 4.5, height = 4)
p
dev.off()
```

## Figure S5D: DSS-induced colitis data
```{r}
dir_colitis <- "../../2_pipeline/colitis/"
trt <- "DSS"
so <- readRDS(file = paste0(dir_colitis, "so_COL_", trt, "_cleaned.rds"))

sigs_ls <- readRDS("../../2_pipeline/spl_clusters/bulk/2020-02-04/signature_slamf6_vs_cxcr6_filtered.rds")
names(sigs_ls) <- c("SLAMF6+_bulk", "CXCR6+_bulk")

## signature scores
so <- AddModuleScore(so, features = sigs_ls, name = "SIGS")
colnames(so@meta.data)[grep("SIGS", colnames(so@meta.data))] <- names(sigs_ls)
plt_df <- data.frame(UMAP_1 = so@reductions$umap@cell.embeddings[,1], 
                     UMAP_2 = so@reductions$umap@cell.embeddings[,2],
                     slamf6_sc = so$`SLAMF6+_sc`,
                     cxcr6_sc = so$`CXCR6+_sc`,
                     slamf6_bulk = so$`SLAMF6+_bulk`,
                     cxcr6_bulk = so$`CXCR6+_bulk`) %>% 
  mutate(slamf6_sc_rank_norm = rank(slamf6_sc) / length(slamf6_sc),
         cxcr6_sc_rank_norm = rank(cxcr6_sc) / length(cxcr6_sc),
         slamf6_bulk_rank_norm = rank(slamf6_bulk) / length(slamf6_bulk),
         cxcr6_bulk_rank_norm = rank(cxcr6_bulk) / length(cxcr6_bulk)) 
p_slamf6_bulk <- plt_df %>% 
  ggplot(.,  aes(x = UMAP_1, y = UMAP_2, color = slamf6_bulk_rank_norm)) + 
  geom_point(size = pt.size) +
  scale_color_gradientn(colours = colorRamps::matlab.like2(10)) +
  labs(color = "Normalized\nsignature\nscore", title = "SLAMF6+ Signature") +
  theme_classic()
p_cxcr6_bulk <- plt_df %>% 
  ggplot(.,  aes(x = UMAP_1, y = UMAP_2, color = cxcr6_bulk_rank_norm)) + 
  geom_point(size = pt.size) +
  scale_color_gradientn(colours = colorRamps::matlab.like2(10)) +
  labs(color = "Normalized\nsignature\nscore", title = "CXCR6+ Signature") +
  theme_classic()

pdf(file = paste0(dir_out, "FigS5D.pdf"), width = 9, height = 4)
grid.arrange(p_slamf6_bulk + theme(legend.position = "none"), 
             p_cxcr6_bulk + theme(legend.position = "none"), 
             nrow = 1, right = get_legend(p_slamf6_bulk))
dev.off()
```


# Figure S6
```{r}
dir_out <- paste0(dir_proj, "3_output/Figure_S6/")
if(!dir.exists(dir_out)) dir.create(dir_out, recursive = T)
```

## Figure S6A: TCR-specificity of tissue Th17 cells by GLIPH2
```{r}
two_beta <- readRDS(file = "../../2_pipeline/TCR/gliph/cells_two_beta.rds")
res <- read.table("../../2_pipeline/TCR/gliph/2021-09-22_ut_cluster.csv", sep = ",", header = T)
res <- res %>% 
  filter((grepl("global", type) & (Fisher_score < 0.05) & (pattern != "single")))  # only keep significant global clusters
length(unique(res$index[grepl("global", res$type)]))

two_beta_df <- Reduce(rbind, two_beta[grep("UT", names(two_beta))]) %>% 
  mutate(CDR3b_1 = sub(":.*", "", CDR3b),
         CDR3b_2 = sub(".*:", "", CDR3b))
two_beta_df <- unique(two_beta_df[,-1])

idx_1 <- which(two_beta_df$CDR3b_1 %in% res$TcRb) # which cells are found in a valid cluster based on the first beta chain
idx_2 <- which(two_beta_df$CDR3b_2 %in% res$TcRb) # which cells are found in a valid cluster based on the second beta chain
idx_1_2 <- intersect(idx_1, idx_2) # both chains are in cluster(s)

res$index_merge <- res$index
for (i in idx_1_2){
  res_idx_1 <- which(as.character(res$TcRb) == as.character(two_beta_df[i,"CDR3b_1"]))
  res_idx_2 <- which(as.character(res$TcRb) == as.character(two_beta_df[i,"CDR3b_2"]))
  cluster_idx_1 <- res$index[res_idx_1[1]]
  cluster_idx_2 <- res$index[res_idx_2[1]]
  cat("original cluster index for", i, ":", cluster_idx_1, cluster_idx_2, "\n")
  new_cluster_idx <- paste0(cluster_idx_1, "_m")
  res$index_merge[res$index == cluster_idx_1] <- new_cluster_idx
  res$index_merge[res$index == cluster_idx_2] <- new_cluster_idx
}

mtx <- res %>% 
  mutate(tissue = sub(".*:", "", Sample),
         treatment = sub(".*?_", "", Sample) %>% sub("_.*", "", .),
         tt = paste(tissue, treatment, sep = "_"),
         index = paste0("Cluster_", index_merge)) %>% 
  group_by(index, tissue) %>% 
  summarise(cluster_size = sum(Freq)) %>% 
  ungroup() %>% 
  spread(key = tissue, value = cluster_size) %>% 
  column_to_rownames(var = "index")
mtx[is.na(mtx)] <- 0
mtx <- mtx[,tissue_vec_ut]
cluster_size_vec <- rowSums(mtx)
summary(cluster_size_vec)
mtx <- mtx[cluster_size_vec > 0, ]
cluster_size_vec <- rowSums(mtx)
summary(cluster_size_vec)
pct_mtx <- mtx / rowSums(mtx) * 100

which_specific <- which((rowSums(mtx > 0)) == 1)
which_not_specific <- setdiff(1:nrow(mtx), which_specific)
idx_row_1 <- which_specific[do.call(order, -cbind(pct_mtx, cluster_size_vec)[which_specific, ])] 
idx_row_2 <- which_not_specific[do.call(order, -cbind(pct_mtx, cluster_size_vec)[which_not_specific,])]
idx_row <- c(idx_row_1, idx_row_2) 
pct_mtx <- pct_mtx[idx_row,]
p <- pheatmap(pct_mtx,
         color = colorRampPalette(c("white", "red"))(100),
         cluster_rows = F,
         cluster_cols = F,
         clustering_method = "complete",
         show_rownames = F,
         silent = F, 
         main = "")
which_col_names <- which(p$gtable$layout$name == "col_names")
p$gtable$grobs[[which_col_names]]$rot = 0
p$gtable$grobs[[which_col_names]]$hjust = 0.5
p$gtable$grobs[[which_col_names]]$vjust = 1
p$gtable$grobs[[which_col_names]]$gp$fontsize = 10

pdf(paste0(dir_out, "FigS6A.pdf"), width = 2.3, height = 5)
grid.arrange(p$gtable)
dev.off()
```


## Figure S6B: UMAPs of EAE tissue clusters
```{r}
library(openxlsx)
clustering_date <- "2020-03-25"
dir_clustering <- paste0(dir_proj, "2_pipeline/clustering/EAE_GFPall_intra/", clustering_date, "/")
dir_spl_cluster <- paste0(dir_proj, "2_pipeline/clustering/EAE_GFPall_SPL/", clustering_date, "/")

tissue_vec <- c("SPL", "MLN", "PP", "SI", "COL", "CNS", "DLN")
### load clustering data
meta <- list()
reductions <- list()
for (tissue in tissue_vec) {
  if (tissue == "SPL") {
    meta[[tissue]]  <- read.table(paste0(dir_spl_cluster, "FILES/meta_data.csv"), header = T, row.names = 1, sep = ",")
    meta[[tissue]] <- meta[[tissue]][,-grep("integrated_snn_res", colnames(meta[[tissue]]))]
    meta[[tissue]]$tissue_cluster <- paste(meta[[tissue]]$tissue, meta[[tissue]]$seurat_clusters, sep = "_")
    reductions[[tissue]] <- readRDS(paste0(dir_spl_cluster, "FILES/reductions.rds"))
  } else if (tissue == "CNS") {
    meta[[tissue]]  <- read.table(paste0(dir_clustering, "FILES/", tissue, "/meta_data_switched.csv"), header = T, row.names = 1, sep = ",")
    meta[[tissue]] <- meta[[tissue]][,-grep("integrated_snn_res", colnames(meta[[tissue]]))]
    meta[[tissue]]$tissue_cluster <- paste(meta[[tissue]]$tissue, meta[[tissue]]$seurat_clusters, sep = "_")
    reductions[[tissue]] <- readRDS(paste0(dir_clustering, "FILES/", tissue, "/reductions.rds"))
  } else {
    meta[[tissue]]  <- read.table(paste0(dir_clustering, "FILES/", tissue, "/meta_data.csv"), header = T, row.names = 1, sep = ",")
    meta[[tissue]] <- meta[[tissue]][,-grep("integrated_snn_res", colnames(meta[[tissue]]))]
    meta[[tissue]]$tissue_cluster <- paste(meta[[tissue]]$tissue, meta[[tissue]]$seurat_clusters, sep = "_")
    reductions[[tissue]] <- readRDS(paste0(dir_clustering, "FILES/", tissue, "/reductions.rds"))
  }
  if (tissue == "MLN") { 
    idx_contminated <- which(reductions[[tissue]]$umap@cell.embeddings[,"UMAP_2"] > 8)
    reductions[[tissue]]$umap@cell.embeddings <- reductions[[tissue]]$umap@cell.embeddings[-idx_contminated,]
    meta[[tissue]] <- meta[[tissue]][-idx_contminated,]
  }
}
all_meta <- Reduce(rbind, meta)
tissue_cluster_vec <- unique(all_meta$tissue_cluster)

## load annotation
cluster_ann_types <- c("Proliferating", "Treg-like", "Effector-like", "Tfh-like", "ISG-high", "Slamf6+", "Cxcr6+", "Other")
cluster_ann <- read.xlsx(paste0(dir_clustering, "cluster_annotation.xlsx"), sheet = "processed")
cluster_ann <- cluster_ann[!duplicated(cluster_ann),]
cluster_ann$Annotation[is.na(cluster_ann$Annotation)] <- "Other"
stopifnot(length(setdiff(cluster_ann$Cluster, tissue_cluster_vec)) == 0) ## check if there are weird cluster in annotation file
if (length(setdiff(tissue_cluster_vec, cluster_ann$Cluster)) > 0) {
  other_clusters <- data.frame(
    Cluster = setdiff(tissue_cluster_vec, cluster_ann$Cluster),
    Annotation = "Other")
  cluster_ann <- rbind(cluster_ann, other_clusters)
}
## set levels
cluster_ann$Annotation <- factor(cluster_ann$Annotation, levels = cluster_ann_types)
cluster_ann <- cluster_ann[order(cluster_ann$Annotation),]


## plot UMAPs
## cluster colors
max_n_cluster <- sapply(strsplit(tissue_cluster_vec, split = "_"), function(x){x[2]}) %>% as.numeric() %>% max() + 1
color_cluster <- brewer.pal(max_n_cluster, "Set1")
names(color_cluster) <- 0:(max_n_cluster - 1)
## cell type colors
color_ann <- brewer.pal(nlevels(cluster_ann$Annotation), "Dark2")
color_ann[nlevels(cluster_ann$Annotation)] <- "grey50"
names(color_ann) <- levels(cluster_ann$Annotation)

p_cluster <- list()
p_gfp <- list()
p_ann <- list()
for (tissue in tissue_vec) {
  cat(tissue, "\n")
  plt_df <- data.frame(
    cluster = factor(meta[[tissue]]$seurat_clusters),
    ex_curr = factor(ifelse(meta[[tissue]]$GFP_positive, "Current Th17", "Ex Th17"), levels = c("Current Th17", "Ex Th17")),
    cluster_ann = factor(cluster_ann$Annotation[match(meta[[tissue]]$tissue_cluster, cluster_ann$Cluster)], levels = levels(cluster_ann$Annotation)),
    UMAP_1 = reductions[[tissue]]$umap@cell.embeddings[,"UMAP_1"],
    UMAP_2 = reductions[[tissue]]$umap@cell.embeddings[,"UMAP_2"]
  )
  p_cluster[[tissue]] <- ggplot(plt_df, aes(x = UMAP_1, y = UMAP_2, color = cluster)) + 
    geom_point(size = 0.1) +
    scale_color_manual(values = color_cluster) +
    guides(color = guide_legend(override.aes = list(size = 5))) +
    labs(x = "", y = "", color = "Cluster", title = tissue) +
    theme_classic() +
    theme(plot.title = element_text(colour = tissue_colors(tissue), hjust = 0, face = "bold", size = 20),
          legend.position = "none",
          axis.ticks = element_blank(), axis.text = element_blank())
  p_ann[[tissue]] <- ggplot(plt_df, aes(x = UMAP_1, y = UMAP_2, color = cluster_ann)) + 
    geom_point(size = 0.1) +
    scale_color_manual(values = color_ann) +
    guides(color = guide_legend(override.aes = list(size = 5))) +
    labs(x = "", y = "", color = "Annotation") +
    theme_classic() +
    theme(legend.position = "none",
          axis.ticks = element_blank(), axis.text = element_blank())
  
  p_gfp[[tissue]] <- ggplot(plt_df, aes(x = UMAP_1, y = UMAP_2, color = ex_curr)) + 
    geom_point(size = 0.1) +
    scale_color_manual(values = gfp_colors()) +
    guides(color = guide_legend(override.aes = list(size = 5))) +
    labs(x = "", y = "", color = "Status") +
    theme_classic() +
    theme(legend.position = "none",
          axis.ticks = element_blank(), axis.text = element_blank())
}

## create legends
color_cluster_pad <- color_cluster
names(color_cluster_pad) <- names(color_cluster)
tmp <- data.frame(x = factor(0:(max_n_cluster-1)), y = 1) %>% 
  ggplot(aes(x, y, color = x)) + 
  scale_color_manual(values = color_cluster_pad) + 
  geom_point() + labs(colour = "Cluster") + 
  theme(legend.text = element_text(size = 15),
        legend.title = element_text(size = 20)) +
  guides(color = guide_legend(override.aes = list(size = 5)))
p_cluster$legend <- as_ggplot(get_legend(tmp)) 

tmp <- data.frame(x = c("Current Th17", "Ex Th17"), y = 1) %>% 
  ggplot(aes(x, y, color = x)) + 
  scale_color_manual(values = gfp_colors()) + 
  geom_point() + theme_bw() + labs(colour = "Status")+ 
  theme(legend.text = element_text(size = 15),
        legend.title = element_text(size = 20)) +
  guides(color = guide_legend(override.aes = list(size = 5)))
p_gfp$legend <- as_ggplot(get_legend(tmp))

tmp <- data.frame(x = cluster_ann$Annotation, y = 1) %>% 
  ggplot(aes(x, y, color = x)) + 
  scale_color_manual(values = color_ann) + 
  geom_point() + theme_bw() + labs(colour = "Annotation") + 
  theme(legend.text = element_text(size = 15),
        legend.title = element_text(size = 20)) +
  guides(color = guide_legend(override.aes = list(size = 5)))
p_ann$legend <- as_ggplot(get_legend(tmp)) 

pdf(file = paste0(dir_out, "FigS6B.pdf"), width = 28, height = 12)
grid.arrange(grobs = c(p_cluster, p_ann, p_gfp), nrow = 3, ncol = 8, 
             left = textGrob(label = "UMAP_2", rot = 90, gp = gpar(fontsize = 20)), 
             bottom = textGrob(label = "UMAP_1", rot = 0, gp = gpar(fontsize = 20)))
dev.off()


```

## Figure S6C: Cluster correlation and annotation plot
```{r}
library(corrplot)
analysis_date <- "2020-05-11"
results <- readRDS(paste0(dir_clustering, "cluster_similarity_", analysis_date, ".rds"))
## prep data
hc_order = results$hierarchical_clustering$order
sim_mat_transformed <- results$cosine_sim[hc_order, hc_order]
diag(sim_mat_transformed) <- NA
tl.col <- color_ann[cluster_ann$Annotation[match(rownames(sim_mat_transformed), cluster_ann$Cluster)]]

pdf(paste0(dir_out, "FigS6C.pdf"), width = 7, height = 7)
corrplot(sim_mat_transformed, type="upper", order="original", method = "circle",
         col=rev(brewer.pal(n=8, name="RdYlBu")),
         mar = c(5, 4, 4, 2) + 0.1,
         tl.col = tl.col, tl.pos = 'td', tl.cex = 0.6,
         diag = T, na.label = " ",
         title = "Cluster Label Color by Annotation")
dev.off()

```



## Figure S6D: Dotplot for cluster marker genes
```{r}
glist_df <- read.xlsx(paste0(dir_clustering, "cluster_annotation.xlsx"), sheet = "genes_processed")
glist_df <- glist_df[glist_df$Annotation %in% unique(cluster_ann$Annotation),] ## only show types that do exist
glist_df$Annotation <- factor(glist_df$Annotation, levels = cluster_ann_types)
glist_df <- glist_df[order(glist_df$Annotation),]

color_glist <- color_ann[glist_df$Annotation]
names(color_glist) <- glist_df$Gene
color_dp_cluster <- color_ann[cluster_ann$Annotation[match(levels(results$dotplot_df$tissue_cluster), cluster_ann$Cluster)]]
color_dp_cluster <- color_dp_cluster[names(color_dp_cluster) != "Other"]

p <- results$dotplot_df %>% 
  mutate(gene = factor(gene, levels = glist_df$Gene)) %>% 
  ggplot() +
  geom_point(aes(x = gene, y = tissue_cluster, size = pct, color = expr)) +
  labs(x = "", y = "") +
  scale_colour_gradient(low = "gray90", high = "black") +
  scale_size(range = c(0, 4), trans = scales::trans_new("square", function(x) x^2, "sqrt")) +
  scale_x_discrete(labels = glist[glist_order]) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, color = color_glist), 
        axis.text.y = element_text(color = color_dp_cluster),
        legend.title = element_text(size = 8),
        legend.box = "horizontal", legend.position="bottom", legend.key.size = unit(.13, "in")) +
  guides(color = guide_colorbar(title = "Scaled Average\nExpression", title.vjust = .8),
         size = guide_legend(title = "Percent Cell\nDetected", title.vjust = .5, ))
p
pdf(paste0(dir_out, "FigS10_C_doplot_EAE_tissue_cluster_markers_", today, ".pdf"), width = 6.5, height = 6)
print(p)
dev.off()
```




## Figure S6E: MA plot for SI_5 vs. SI_other.
```{r}
date_de <- "2020-05-18"
de_si <- readRDS(paste0(dir_proj, "2_pipeline/differential_expression/Cluster_EAE_GFPall_intra/", date_de, "/SI_results.rds"))

library(ggrepel)
thresh_log2fc <- log2(1.5)
thresh_fdr <- 0.05
plt_df <- de_si$C_5 %>%
  rownames_to_column(var = "gene") %>%
  filter(!(grepl("Rps", gene) | grepl("Rpl", gene))) %>% 
  mutate(is_de = abs(logFC) > thresh_log2fc & FDR < thresh_fdr) %>%
  mutate(de_direction = ifelse(is_de & logFC > 0, "up", ifelse(is_de & logFC < 0, "down", "ns"))) %>%
  mutate(label = ifelse( (is_de & (logFC < -2 | logFC > 1.7 | logCPM > 11)), gene, NA))

p <- ggplot(plt_df, aes(x = logCPM, y = logFC, color = de_direction)) +
  geom_point() +
  geom_label_repel(aes(label = label)) +
  scale_color_manual(values = c("up" = "red", "down" = "blue", "ns" = "black")) +
  labs(x = "Average Log2 CPM", y = "Log2 Fold Change (SI_5 vs. others)") +
  theme_bw() +
  theme(legend.position = "none")

p
pdf(paste0(dir_out, "FigS6E.pdf"), width = 7, height = 5)
p
dev.off()

```

## Figure S6F: Enrichment for SI_5 DE genes
```{r}
library(fgsea)
source(paste0(dir_proj, "1_code/enrichment/utils.R"))

go_bp <- read_parse_db(paste0(dir_proj, "0_data/other/GO_Biological_Process_2018.txt"))
go_bp <- human_to_mouse(go_bp) %>% lapply(., toupper)
kegg <- read_parse_db(paste0(dir_proj, "0_data/other/KEGG_2019_Mouse.txt"))

gsea <- list()
gsea$go_bp <- test_genesets(db = go_bp, res = de_si5$C_5, reverse = FALSE)
gsea$kegg <- test_genesets(db = kegg, res = de_si5$C_5, reverse = FALSE)
gsea$go_bp$database <- "GO"
gsea$kegg$database <- "KEGG"
gsea_combined <- rbind(gsea$go_bp, gsea$kegg)

original_terms <- c("regulation of actin cytoskeleton",
                    "leukocyte transendothelial migration", 
                    "focal adhesion", 
                    "regulation of t cell activation (go:0050863)",
                    "tight junction",
                    "cell adhesion molecules (cams)",
                    "integrin-mediated signaling pathway (go:0007229)")
parsed_terms <- c("Regulation of actin cytoskeleton (KEGG)",
                    "Leukocyte transendothelial migration (KEGG)", 
                    "Focal adhesion (KEGG)", 
                    "Regulation of t cell activation (GO)",
                    "Tight junction (KEGG)",
                    "Cell adhesion molecules (KEGG)",
                    "Integrin-mediated signaling pathway (GO)")

gsea_selected <- gsea_combined[match(original_terms, tolower(gsea_combined$pathway)),]
gsea_selected$pathway_parsed <- parsed_terms

## plot
library(stringr)
p <- gsea_selected %>% 
  arrange(NES) %>% 
  mutate(pathway_name = str_wrap(parsed_terms, width = 30)) %>% 
  mutate(pathway_name = factor(pathway_name, levels = pathway_name)) %>% 
  ggplot() +
  geom_bar(aes(x = pathway_name, y = NES), stat = "identity", width = 0.6, color = "black", fill = "white") +
  geom_text(aes(x = pathway_name, y = NES, label = paste0("p=", formatC(pval, digits = 1, format = "E"))), hjust = 1.1) +
  labs(x = "", y = "NES", fill = "") +
  coord_flip() +
  theme_bw()
p
pdf(paste0(dir_out, "FigS6F.pdf"), width = 5, height = 5)
p 
dev.off()
```

## Figure S6G: Number of cells sharing TCR with CNS in each tissue
```{r}
## clonotypes
clonotyping_type_date <- "dominant_2020-04-02"
clty_all <- read.csv(paste0(dir_proj, "2_pipeline/TCR/clonotype_assignment_", clonotyping_type_date, ".csv"), row.names = 1)

## merge clonotype info to meta data
all_meta$clonotype_id <- clty_all[rownames(all_meta),"clonotype_id"]
all_meta$pooled_clonotype_id <- paste(all_meta$treatment, all_meta$batch, all_meta$mouse, all_meta$clonotype_id, sep = "_")
all_meta$sample <- paste(all_meta$batch, all_meta$mouse, sep = "_") %>% factor()
all_meta$tissue_cluster <- factor(paste(all_meta$tissue, all_meta$seurat_clusters, sep = "_"))

meta_noNA <- all_meta[!is.na(all_meta$clonotype_id),] %>% droplevels()
stopifnot(all(meta_noNA$treatment == "EAE")) ## check if treatment is correct (should be EAE cells only)

tissue_vec_sub <- c("MLN", "PP", "SI", "COL")
cl_cns <- unique(meta_noNA$pooled_clonotype_id[meta_noNA$tissue == "CNS"])

plt_df <- meta_noNA %>% 
  filter(tissue %in% tissue_vec_sub) %>% 
  group_by(tissue) %>% 
  summarise(n_shared = sum(pooled_clonotype_id %in% cl_cns),
            n_total = length(pooled_clonotype_id)) %>% 
  mutate(pct_shared = 100 * n_shared / n_total,
         tissue = factor(tissue, levels = tissue_vec_sub))

p <- ggplot(plt_df, aes(x = tissue, y = n_shared, fill = tissue)) +
  geom_bar(stat = "identity", width = 0.8) +
  geom_text(aes(y = n_shared/2, label = paste0(n_shared, "\n(", sprintf("%.2f", pct_shared), "%)")), size = 4.5) +
  scale_fill_manual(values = tissue_colors(tissue_vec)) +
  labs(x = "", y = "Number of Cells") +
  theme_bw() +
  theme(legend.position = "none",
        axis.text = element_text(size = 12, color = "black"),
        axis.title.y = element_text(size = 15, color = "black"))
p
pdf(paste0(dir_out, "FigS6G.pdf"), width = 4, height = 4)
p
dev.off()
```

## Figure S6H: SI_5 signature scores
```{r}
## find cells sharing TCR with CNS
meta_noNA <- meta_noNA %>% group_by(pooled_clonotype_id) %>% 
  mutate(shared_with_CNS = any(tissue == "CNS"))
cells_shared_with_CNS <- sapply(tissue_vec, function(tissue) {
  meta_noNA$cell[meta_noNA$tissue == tissue & meta_noNA$shared_with_CNS] %>% as.character()
})

## signatures
sig_si5 <- rownames(de_si5$C_5)[de_si5$C_5$logFC > log2(1.5) & de_si5$C_5$FDR < 0.05]
sigs <- list(sig_si5)

## compute scores
tissue_vec_sub <- c("MLN", "PP", "SI", "COL")
so <- so_all[,so_all$treatment == "EAE"]
so <- AddModuleScore(so, features = sigs, name = "SI_5")
so$shared_with_CNS <- meta_noNA$shared_with_CNS[match(colnames(so), meta_noNA$cell)]
so$shared_with_CNS[is.na(so$shared_with_CNS)] <- "No TCR"
so$shared_with_CNS <- factor(so$shared_with_CNS, levels = c("TRUE", "FALSE", "No TCR"), labels = c("Shared", "Not shared", "No TCR"))

sig_scores <- so@meta.data %>%
  rownames_to_column(var = "cell_name") %>%
  select(cell_name, score = SI_51, tissue, shared_with_CNS)

## box plot
p_score_box <- sig_scores %>% 
  filter(shared_with_CNS %in% c("Shared", "Not shared")) %>% 
  filter(tissue %in% tissue_vec_sub) %>% 
  mutate(tissue = factor(tissue, levels = tissue_vec_sub)) %>% 
  ggplot(aes(x = shared_with_CNS, y = score)) +
  geom_boxplot(aes(fill = tissue), show.legend = F, width = 0.5) +
  facet_wrap(~tissue, nrow = 1) +
  ylim(-1, 1.6) +
  geom_signif(test="t.test", comparisons = list(c("Shared", "Not shared")), map_signif_level = T, margin_top = 0.1) +
  scale_fill_manual(values = tissue_colors()) +
  labs(x = "", y = "SI_5 Signature Score", fill = "") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
p_score_box

pdf(paste0(dir_out, "FigS6H.pdf"), width = 5, height = 5)
p_score_box
dev.off()
```

# Figure S7
```{r}
dir_out <- paste0(dir_proj, "3_output/Figure_S7/")
if(!dir.exists(dir_out)) dir.create(dir_out, recursive = T)
```

## Figure S7C: Violin plot compare proliferation signature in SPL EAE cluster 0 and 1
```{r}
## load data
date_cluster <- "2020-03-25"
meta_spl <- read.table(paste0(dir_proj, "2_pipeline/clustering/EAE_GFPall_SPL/", date_cluster, "/FILES/meta_data.csv"), 
                   sep = ",", header = T, row.names = 1)
meta_spl <- meta_spl[meta_spl$seurat_clusters %in% c(0, 1),]
so_sub <- so_all[,rownames(meta_spl)]
so_sub$tissue_cluster <- paste(so_sub$tissue, so_sub$seurat_clusters, sep = "_")

## read and parse proliferating signature from our UT GFP+ inter tissue clustering
de <- readRDS(paste0(dir_proj, "2_pipeline/differential_expression/Cluster_UT_GFPpos_inter/2020-04-16/results.rds"))
sig <- de[["7"]] %>% rownames_to_column(var = "gene") %>% 
  filter(FDR < 0.05 & logFC > log2(1.5)) %>% 
  arrange(FDR) %>% select(gene) %>% unlist %>% as.character()
is_detectd <- Matrix::rowSums(so[["RNA"]]@counts[sig,]) > 0 

## compute score 
so_sub <- AddModuleScore(so_sub, list(sig), name = "SIG")
colnames(so_sub@meta.data)[colnames(so_sub@meta.data) == "SIG1"] <- "Proliferating_all"

## make violin plot
# select cells to show as jitter points
set.seed(1)
so_sub$show_point <- runif(ncol(so_sub)) < 0.50
spl_cols <- SPL_EAE_cluster_0_1_colors()

plt_df <- so_sub@meta.data[, c("Proliferating_all", "tissue_cluster", "show_point")]
w_test_out <- wilcox.test(x = plt_df$Proliferating_all[plt_df$tissue_cluster == "SPL_0"],
                     y = plt_df$Proliferating_all[plt_df$tissue_cluster == "SPL_1"],
                     paired = F, var.equal = F)
plt_df_sub <- plt_df %>% 
  filter((Proliferating_all > quantile(Proliferating_all, 0.001)) & (Proliferating_all < quantile(Proliferating_all, 0.999)))
p <- ggplot(data = plt_df_sub, 
            mapping = aes(x = tissue_cluster, y = Proliferating_all, fill = tissue_cluster)) +
  geom_violin() +
  geom_jitter(aes(shape = show_point), size = 0.01) + 
  geom_boxplot(width = 0.1, fill = NA, color = "white", outlier.shape = NA) +
  geom_signif(xmin = "SPL_0", xmax = "SPL_1", y_position = 1.1 * max(plt_df_sub$Proliferating_all), 
              vjust = 0.6, textsize = 10,
              annotations = map_pval(w_test_out$p.value)) +
  scale_shape_manual(values = c(NA, 19), guide = FALSE) +
  scale_fill_manual(values = spl_cols) + 
  labs(x = "", y = "Signature Score") +
  ggtitle("Proliferation") + 
  theme_classic() +
  theme(legend.position = "none",
        axis.title = element_text(size = 15, color = "black"),
        axis.text = element_text(size = 12, color = "black"),
        plot.title = element_text(hjust = 0.5, face = "bold"))
p
pdf(paste0(dir_out, "FigS7C.pdf"), width = 3, height = 4)
p
dev.off()
```

## Figure S7J: Il23 signature scores
```{r}
## parse signatures
fname <- "0_data/gene_lists/signatures/IL23_signatures.xlsx"
sigs_to_plot <- getSheetNames(fname)
sigs_parsed <- lapply(sigs_to_plot, function(x){
  tmp <- read.xlsx(fname, sheet = x, rowNames = F, colNames = F)[,1]
  if (grepl("human", x)) {
    genes <- human_to_mouse(tmp)
  } else {
    genes <- rownames(so)[match(toupper(tmp), toupper(rownames(so)))] %>% setdiff(., NA)
  }
  genes
})
names(sigs_parsed) <- sigs_to_plot

## compute scores
so <- AddModuleScore(so, sigs_parsed, name = "SIG")
sig_score_df <- so@meta.data[, grep("SIG", colnames(so@meta.data))]
colnames(sig_score_df) <- names(sigs_parsed)

## make plot
plt_df <- data.frame(
  tissue_cluster = meta_spl$tissue_cluster, 
  score = sig_scores$`SGK1-IL23`,
  show_point = show_point
)
w_test_out <- wilcox.test(x = plt_df$score[plt_df$tissue_cluster == "SPL_0"],
                   y = plt_df$score[plt_df$tissue_cluster == "SPL_1"],
                   paired = F, var.equal = F)
plt_df_sub <- plt_df %>% 
  filter((score > quantile(score, 0.001)) & (score < quantile(score, 0.999)))
p <- ggplot(data = plt_df_sub, 
            mapping = aes(x = tissue_cluster, y = score, fill = tissue_cluster)) +
  geom_violin() +
  geom_jitter(aes(shape = show_point), size = 0.01) + 
  geom_boxplot(width = 0.1, fill = NA, color = "white", outlier.shape = NA) +
  geom_signif(xmin = "SPL_0", xmax = "SPL_1", y_position = 1.1 * max(plt_df_sub$score), 
              vjust = 0.6, textsize = 10,
              annotations = map_pval(t_test_out$p.value)) +
  scale_shape_manual(values = c(NA, 19), w_test_out = FALSE) +
  scale_fill_manual(values = SPL_EAE_cluster_0_1_colors()) + 
  labs(x = "", y = "Signature Score", title = "IL-23 signature") +
  theme_classic() +
  theme(legend.position = "none",
        axis.title = element_text(size = 15, color = "black"),
        axis.text = element_text(size = 12, color = "black"),
        plot.title = element_text(hjust = 0.5, face = "bold"))

pdf(paste0(dir_out, "FigS7J.pdf"), width = 3, height = 4)
p
dev.off()
```

## Figure S7K: Il23 signatures in ATAC-seq data
See ../atac/SLAMF6_vs_CXCR6.Rmd

## Figure S7M: Il12b and Il23a expression (Wheeler & Quintana data)
```{r}
dir_wheeler <- "../../2_pipeline/cxcl16/2020-02-26/"
so <- readRDS(paste0(dir_wheeler, "so_combined.rds"))

### annotate clusters
## quantile normalization
library(preprocessCore)
annot_names <- colnames(so@meta.data)[grep("Cluster_", colnames(so@meta.data))]
scores_q <- normalize.quantiles(as.matrix(so@meta.data[,annot_names]))
so$annotation_score <- factor(annot_names[apply(scores_q, 1, which.max)] %>% sub("Cluster_", "C", .), levels = sub("Cluster_", "C", annot_names))
sample_info <- c(
"GSM3717013", "Priming_replicate1",
"GSM3717014", "Priming_replicate2",
"GSM3717015", "Priming_replicate3",
"GSM3717016", "CFA_replicate1",
"GSM3717017", "Acute_replicate1",
"GSM3717018", "Acute_replicate2",
"GSM3717019", "Acute_replicate3",
"GSM3717020", "Naive_replicate1",
"GSM3717021", "Naive_replicate2",
"GSM3717022", "Remitting_replicate1",
"GSM3717023", "Remitting_replicate2",
"GSM3717024", "Remitting_replicate3",
"GSM3717025", "CFA_replicate2",
"GSM3717026", "CFA_replicate3",
"GSM3717027", "Naive_replicate3",
"GSM4002707", "Acute_replicate4",
"GSM4002708", "Acute_replicate5",
"GSM4002709", "Acute_replicate6",
"GSM4002710", "Naive_replicate4",
"GSM4002711", "Remitting_replicate4",
"GSM4002712", "Remitting_replicate5",
"GSM4002713", "Remitting_replicate6",
"GSM4002714", "Naive_replicate5",
"GSM4002715", "Priming_replicate4",
"GSM4002716", "Priming_replicate5",
"GSM4002717", "Priming_replicate6",
"GSM4002718", "Naive_replicate6"
) %>% matrix(ncol = 2, byrow = T) %>% data.frame() %>% 
  select(GSM = X1, sample_name = X2) %>% 
  mutate(treatment = gsub( "_.*$", "",  sample_name))

so$sample_name <- sample_info$sample_name[match(so$orig.ident, sample_info$GSM)]
so$condition <-  factor(gsub( "_.*$", "",  so$sample_name), levels = c("Naive", "Priming", "CFA", "Acute", "Remitting"))

plt_df <- data.frame(
  Il23a = so[["RNA"]]@data["Il23a",],
  Il12b = so[["RNA"]]@data["Il12b",],
  condition = so$condition,
  cluster = so$annotation_score,
  subject = so$sample_name
) %>% filter(condition == "Priming") %>% 
  group_by(cluster, subject, .drop = FALSE) %>% 
  summarise(pct_Il23a = 100 * mean(Il23a > 0),
         pct_Il12b = 100 * mean(Il12b > 0)) %>% 
  ungroup() %>% 
  group_by(cluster, .drop = FALSE) %>% 
  summarise(avg_pct_Il23a = mean(pct_Il23a, na.rm = T),
            hi_pct_Il23a = avg_pct_Il23a + sd(pct_Il23a, na.rm = T) / sqrt(sum(!is.na(pct_Il23a))),
            lo_pct_Il23a = pmax(avg_pct_Il23a - sd(pct_Il23a, na.rm = T) / sqrt(sum(!is.na(pct_Il23a))), 0),
            avg_pct_Il12b = mean(pct_Il12b, na.rm = T),
            hi_pct_Il12b = avg_pct_Il12b + sd(pct_Il12b, na.rm = T) / sqrt(sum(!is.na(pct_Il12b))),
            lo_pct_Il12b = pmax(avg_pct_Il12b - sd(pct_Il12b, na.rm = T) / sqrt(sum(!is.na(pct_Il12b))), 0))

p_il23a <- ggplot(plt_df) +
  geom_bar(aes(x = cluster, y = avg_pct_Il23a), 
           stat = "identity", width = 0.6, color = "black", fill = "#808080", size = 0.5) +
  geom_errorbar(aes(x = cluster, ymin = lo_pct_Il23a, ymax = hi_pct_Il23a), width = 0.3) +
  scale_y_continuous(expand = c(0, 0)) +
  labs(x = "", y = "Percent of cells") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 8, color = "black"),
        plot.margin = unit(c(0.5, 0.5, 0.5, 0.5), "in"))

p_il12b <- ggplot(plt_df) +
  geom_bar(aes(x = cluster, y = avg_pct_Il12b), 
           stat = "identity", width = 0.6, color = "black", fill = "#808080", size = 0.5) +
  geom_errorbar(aes(x = cluster, ymin = lo_pct_Il12b, ymax = hi_pct_Il12b), width = 0.3) +
  scale_y_continuous(expand = c(0, 0)) +
  labs(x = "", y = "Percent of cells") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 8, color = "black"),
        plot.margin = unit(c(0.5, 0.5, 0.5, 0.5), "in"))

pdf(paste0(dir_out, "Fig7M.pdf"), width = 7.5, height = 5.5)
p_il12b
p_il23a
dev.off()
```

## Figure S7N: Il23r KO vs. WT in CXCR6+ cells
```{r}
thresh_padj <- 0.05
date_de <- "2021-06-09"
res_edger_df <- list()
e <- "As308_only"
cell_type <- "CXCR6+"

res_edger <- readRDS(file = paste0(dir_il23rko, "de_edger_", cell_type, "_", e, "_", date_de, ".rds"))

plt_df <- as.data.frame(res_edger) %>%
    rownames_to_column(var = "gene") %>%
    filter(!is.na(PValue)) %>%
    mutate(up_in = ifelse(FDR >= thresh_padj, "NS", ifelse(logFC < 0, "WT", "KO")) %>% factor(., levels = c("KO", "WT", "NS")))
expr <- assay(vsd_cxcr6_308)[plt_df$gene[plt_df$up_in != "NS"],] %>% t %>% scale %>% t
hc_row <- hclust(dist(expr), method = "complete")
expr <- expr[hc_row$order,]
## heatmap
ann_col <- data.frame(
  Condition = toupper(as.character(colData(dds_cxcr6_308)[,"condition"]))
)
rownames(ann_col) <- colnames(dds_cxcr6_308)
ann_colors <- list(Condition = c("WT" = "black", "KO" = rgb(76, 169, 95, maxColorValue = 255)))
p <- pheatmap(expr,
         cluster_cols = F, cluster_rows = F,
         color = colorRampPalette(c("blue", "white", "red"))(100),
         breaks = seq(-2, 2, length.out = 101),
         annotation_col = ann_col,
         annotation_colors = ann_colors,
         show_rownames = T,
         show_colnames = F,
         main = "")
which_row_names <- which(p$gtable$layout$name == "row_names")
p$gtable$grobs[[which_row_names]]$gp$fontsize = 3
p$gtable$grobs[[which_row_names]]$gp$font = 3L

pdf(paste0(dir_out, "FigS7N.pdf"), width = 5, height = 7)
grid.arrange(p$gtable)
dev.off()
```

## Figure S7O: GSEA of Th17 pathogenicity signatures in Il23r KO vs. WT CXCR6+ cells
GSEA barplot.
```{r}
date_fgsea <- "2021-07-29"
test_out <- readRDS(file = paste0(dir_il23rko, "fGSEA_", cell_type, "_", date_fgsea, ".rds"))
sigs_to_plot <- c(
  'Th17_Th1_like_memory-minus' = 'Th17/Th1-like memory\n(minus)',
  'Th17_Th1_like_effector_CNS-minus' = 'Th17/Th1-like effector in CNS\n(minus)',
  'Th17_Th1_like_effector-minus' = 'Th17/Th1-like effector\n(minus)',
  'Th17_self_renewing-plus' = 'Th17 self-renewing',
  'Pathogenic_Th17' = 'Pathogenic Th17',
  '_x0009_HALLMARK_INTERFERON_GAMMA_RESP' = 'HALLMARK interferon\ngamma response'
)
## make bar plot
p <- test_out %>% 
  filter(pathway %in% names(sigs_to_plot)) %>%
  mutate(signature = sigs_to_plot[pathway]) %>%
  mutate(enriched_in = ifelse(NES > 0, "KO", "WT")) %>%
  ggplot(aes(x = reorder(signature, NES), y = NES)) +
  geom_col(aes(fill = enriched_in), width = 0.7) +
  geom_hline(yintercept = 0, color = "black") +
  scale_fill_manual(values = c("KO" = rgb(76, 169, 95, maxColorValue = 255), 
                               "WT" = "black")) +
  coord_flip() +
  labs(x="", y="Normalized Enrichment Score", fill = "Enriched in") +
  theme_bw() +
  theme(legend.title = element_text(size = 12),
        legend.text = element_text(size = 12),
        legend.box.margin=margin(-10,-10,-10,-10),
        axis.title = element_text(size = 12),
        axis.text = element_text(size = 10, color = "black"))
p
pdf(paste0(dir_out, "FigS7O.pdf"), width = 6, height = 4)
print(p)
dev.off()
```
